---
title: "ETC5512: Wild Caught Data"
subtitle: "Ethical use of data"
author: "Lauren Kennedy"
email: "ETC5512.Clayton-x@monash.edu"
date: "Week 3"
length: "120 minutes"
department: "Department of Econometrics and Business Statistics"
titlebgimg: "images/bg-03.png"
output:
  xaringan::moon_reader:
    css:
      - ninjutsu 
      - "assets/font-awesome-all.css"
      - "assets/tachyons-addon.css"
      - "assets/animate.css"
      - "assets/fira-code.css"
      - "assets/boxes.css"
      - "assets/table.css"
      - "assets/styles.css"
      - "assets/monash-brand.css"
      - "assets/monash-fonts.css"
      - "assets/slide-types.css"
      - "assets/custom.css"
      - "assets/emi.css"
    self_contained: false 
    seal: false 
    chakra: 'lib/remark-latest.min.js'
    includes:
      in_header: "assets/head.html"
    lib_dir: lib
    mathjax: "lib/mathjax-local/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    nature:
      highlightStyle: magula
      highlightLanguage: r 
      highlightLines: true
      highlightSpans: false 
      countIncrementalSlides: false
      slideNumberFormat: '%current%/%total%'
      navigation:
        scroll: false 
        touch: true
        click: false
      ratio: '16:9'
---

```{r, include = FALSE}
current_file <- knitr::current_input()
basename <- gsub(".Rmd$", "", current_file)

knitr::opts_chunk$set(
  fig.path = sprintf("images/%s/", basename),
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  fig.retina = 3,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  cache.path = "cache/"
)
```

```{r titleslide, child="assets/titleslide.Rmd"}
```
---
class: middle center

<iframe width="560" height="315" src="https://www.youtube.com/embed/yg16u_bzjPE" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>
---


class: monash-bg-blue middle tr 

.pa5[
.f-headline[**.white[What would you do? How did you decide what to do?**]]

]

---

class: middle center

<iframe width="560" height="315" src="https://www.youtube.com/embed/yg16u_bzjPE?start=37" frameborder="0" allow="accelerometer; autoplay; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

---

#Why consider ethical data practices?

* Data is increasingly accessible and available
  * Can use for understanding our world
  * Can use to make decisions (data driven decision making)
  * Can use to inform policy decisions
--

* Research ethics is often discussed in experimental settings
  * The fundamental ethical concerns do not change
--

* Data has an increasing impact on our lives
  * Data for good initiatives
  * Unethical uses of data
  * Open data increases the reach of data but risks privacy

---


# Ethical Perspectives

* Virtue Based Ethics (Plato, Aristotle, Confucius, Mencius)
  * Act in a way to be a better person (culturally based)
  * The same aim can lead to conflicting actions
--

* Deontological Ethics(Immanuel Kant)
  * Universal ethical laws
  * Rules should apply to everyone
  * Preserve human autonomy and dignity-
--

* Consequential ethics and utilitarianism (John Stuart Mill)
  * Consider the outcomes rather than the intentions
  * Argue some unethical behavior okay if justified by the return


.footnote[
O'Keefe, K., & Brien, D. O. (2018). Ethical data and information management: concepts, tools and methods. Kogan Page Publishers, chapter 2
]

---

# Ethical Perspectives
  
* Common good and justice based ethics
  * Action should contribute to some greater good
--

* Shareholder/stockholder theory (Milton Friedman)
  * Increase value to shareholders by maximizing profits
--
  
* Stakeholder theory
  * Organization responsible to stakeholders
  * Employees, stockholders, society, environment

.footnote[
O'Keefe, K., & Brien, D. O. (2018). Ethical data and information management: concepts, tools and methods. Kogan Page Publishers, chapter 2
]

---


# Applying different ethical practices

.question-box.w-100[
A large company surveys all of their current employees, measuring demographics and
personality factors. They hope to identify key personality factors that correspond with
a successful time with the company. Their hope is to use this data to identify which
prospective employees they should hire. 
]

--

.w-60[
.fr.animated.bounceIn.f1.small-caps[Stakeholder theory]

]

--

.w-60[
.fr.animated.bounceIn.f1.small-caps[Shareholder theory]

]

--

.w-60[
.fr.animated.bounceIn.f1.small-caps[Deontological Ethics]

]
---
class: middle center

# How do we decide what is ethical?

<iframe width="560" height="315" src="https://www.youtube.com/embed/afwK2CVpc9E" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

---
# Belmont report (1979)

* Identifies what constitutes as research
--

* Basic Ethical Principles
--

* Application to demonstrate the use of these principles

.footnote[https://www.hhs.gov/ohrp/regulations-and-policy/belmont-report/read-the-belmont-report/index.html#xbound

]

---
# What constitutes as research?

.blockquote[
**Research**  
"an activity designed to test an hypothesis, permit conclusions to be drawn, and thereby to develop or contribute to generalizable knowledge
"
]
--
.blockquote[
**Practice**  
"interventions that are designed solely to enhance the well-being of an individual patient or client and that have a reasonable expectation of success"
]

.footnote[https://www.hhs.gov/ohrp/regulations-and-policy/belmont-report/read-the-belmont-report/index.html#xbound

]

---
#Basic Ethical Principles
.info-box.w-50[
Autonomy is about being able to deliberate and make personal goals, and then act upon them
]

1. Respect for Persons
 * Acknowledge and respect autonomy
 * Protect those with diminished autonomy
--
 
2. Beneficence
 * Do not cause harm
 * Minimize possible harms and maximise possible benefits
--
 
3. Justice
  * Who ought to receive the benefits of research and bear its burdens?


.footnote[https://www.hhs.gov/ohrp/regulations-and-policy/belmont-report/read-the-belmont-report/index.html#xbound

]

---
# Applications

1. Informed consent
  * Participants have a right to understand what is going to happen
  * Application of respect for persons
  
--

2. Assessment of Risks and Benefits
  * Are the risks justified?
  * Is the study well designed?
  * Are there alternative lower risk designs possible?
  * Are potential benefits and risks communicated to participants?
  * Application of beneficence
  
.footnote[https://www.hhs.gov/ohrp/regulations-and-policy/belmont-report/read-the-belmont-report/index.html#xbound

]

---
# Applications

3. Selection of Subjects
  * Fairly offer research participation to all eligible 
  * Some subjects preferred to as best able to bear risk (e.g., adults)
  * "injustice arises from social, racial, sexual and cultural biases institutionalized in society"
  * Careful not to overburden vulnerable subjects
  
.footnote[https://www.hhs.gov/ohrp/regulations-and-policy/belmont-report/read-the-belmont-report/index.html#xbound
]

---
# Australia- National Statement on ethical conduct in research involving humans

.info-box.w60[Human research is conducted with or about people, or their data or tissue.]

* Governs how to ethically do research involving humans in Australia
--

* Discusses Risk and Benefit, Consent
--

* Discusses ethical considerations for specific participants (e.g., children, pregnant women,
Aboriginal and Torres Strait Islander Peoples)
--

* Discusses when and how ethical review can be conducted
--

* Discusses Risk and the level of review
  * Neglible risk - no foreseeable risk other than inconvenience
  * Low risk - only risk is discomfort
  * Potential risks must be balanced with potential benefit
  
.footnote[https://www.nhmrc.gov.au/about-us/publications/national-statement-ethical-conduct-human-research-2007-updated-2018#block-views-block-file-attachments-content-block-1
]
---
# Who decides what is ethical?
* Human Research Ethics Committee (HREC) in Australia. Can be small differences in other countries so be mindful! (e.g., IRB in US)
--

* In Australia, all research with more than low risk needs to be reviewed by a HREC
--

* Low risk studies might be reviewed under different levels
--

* Studies with negligible risk might be exempt under certain conditions

---

.question-box.w-100[
**Who is on a HREC?**
Sections 5.1.29 = 5.1.30
* Minimum 8 
 * even numbers of men and women
 * 1/3 outside of the institution
* Chairperson
* Two lay people (do not do medical, scientifc, legal or academic work)
* At least one person with knowledge/experience in treatment/professional care (e.g., a nurse)
* One person who is in a pastoral care role (Aboriginal elder, minister)
* One lawyer
* Two people with current research experience
]
---
class: monash-bg-blue middle tr 

.pa5[
.f-headline[**.white[Why do you think the HREC requires such a diverse group members?**]]

]

---
# General Data Protection Laws

* Passed in 2018 to provide strong protection on the collections, use and management of data
--

* Rights of the individual
  * Information and access - you can access your data and see how it is processed
  * Right to request erasure of data
--

* Companies can now be held accountable
--

* In Australia, need to comply if you offer goods and services in EU, collect data from EU individuals or are established in the EU

.footnote[https://gdpr-info.eu/]

---
# Balancing Risk and Benefit

.info-box.w-100[
**National Statement on ehtical conduct in research involving humans**
"risk is a potential for harm, discomfort or
inconvenience (discussed below). It involves:
* the likelihood that a harm (or discomfort
or inconvenience) will occur; and
* the severity of the harm, including its
consequences."
]
--
Risks should be:
- Identified/Assessed
--

- Minimized
--

- Justified - does the potential benefit outweigh the potential risk?
--

- Monitored and managed if the research is approved

---
# Your turn
.question-box.w-100[
A study investigates whether there has been an increased occurrence in loneliness during the COVID-19 lockdowns in Melbourne, Australia. The research method uses a survey administered during lockdown and non-lockdown periods to measure participant loneliness. What is one potential risk? How could this risk be minimized/managed/monitored? Are there benefits that justify this risk?
]

--

* **RISK:** Respondents reports extreme psychological stress
--

* **BENEFIT:** Understanding loneliness patterns can assist in providing preventitive and other mental health services
--

* **MITIGATION/MANAGEMENT:** Respondents are given a list of freely available mental health services to use if needed at the end of the study.

---
# Consent

.info-box.w-100[
**Belmont Report**
* Information
* Comprehension
* Voluntary (without cohersion)
]
--
.info-box.w-100[
**National Statement on ethical conduct in research involving humans**
"...consent should be a voluntary choice, and should
be based on sufficient information and adequate
understanding of both the proposed research and
the implications of participation in it."
]

---
# Your turn
.question-box.w-100[
You are working on a study involving high schoolers aged between 15 and 17. The study aims to investigate the impact of a learning intervention on scores in maths subjects. The students are told they will be randomly assigned to either an intervention or a control condition of maths exercises, and the researchers request access to their future scores as part of the study. Students will be compensated $100 for participation. What should be considered to obtain consent?
]

--

* Do teenagers or their parents consent?
--

* Is $100 a reasonable compensation or could it coerse students into participating
--

* Are alternatives to participating presented to students?
--

* Do participants sacrifice learning time to participate?

---
# Justice, data sources and representation

* From the Belmont report we see that justice relates to the risks/benefits of the study at the same probability for all participants 
--

* They also note the need to be careful not to exhaust vulnerable populations by overstudying. 
--

* Big datasets often have challenges of representation - not everyone in the population has an equal chance of representation
--

* This can be one cause of biased predictions and unfair algorithms - the algorithms are less accurate for particular subsets of the population
--

* However, much of big data doesn't cost the participant to give - it's created through scraping etc, and the lack of representation reflects sociological and cultural power imbalances

---
#Your turn

.question-box.w-100[
A large company surveys all of their current employees, measuring demographics and
personality factors. They hope to identify key personality factors that correspond with
a successful time with the company. Their hope is to use this data to identify which
prospective employees they should hire. 
]

* Are there any concerns with the data source?
--

* Using existing employee data reinforces any previous discriminatory hiring patterns

.info.box.w-100[Counterfactual: What would have happened if we hired people who were different to those previously hired?]



--

---
# Focussing on statistical and data practices

* So far we've focused on ethical and data practice in research generally.
--

* The basic principles apply to ALL research, but many of the examples focus on experimental research 
--

* How can we translate this principles to a more statistical/data focus?

---

# Ethical guidelines for statisticians

.footnote[https://www.amstat.org/asa/files/pdfs/EthicalGuidelines.pdf]

1. Professional Integrity and Accountability
--

2. Integrity of data and method
--

3. Responsibilities to Science/Public/Funder/Client 
--

4. Responsibilities to Research Subjects
--

5. Responsibilities to Research Team Colleagues
--

6. Responsibilities to Other Statisticians or Statistics Practitioners
--

7. Responsibilities Regarding Allegations of Misconduct

---

### Alternative view

.footnote[Gelman (2018). Ethics in statistical practice and communication: Five recommendations. Significance]

1. Prioritize open data and methods
--

2. Be clear about the information and assumptions that go into statistical methods
--

3. Respect for data
--

4. Publication of criticisms 
--

5. Respect the limitations of statistics


---
# Yet another perspective

.footnote[https://deon.drivendata.org/]

* Much of what we spent the first have of the lecture discussing surrounds data collection
--

* Identify other areas including:
 * Storage 
 * Analysis 
 * Modeling
 * Deployment
--

* These areas are less likely to be mentioned, but are equally important. 

---
# How to be an ethical data scientist

1. Firstly ensure that you are following the ethical frameworks of the company you are in (e.g., I work for Monash so I ensure that my research conforms to Australian standards through Monash procedures)
--

2. Ensure you understand the different areas where unethical thinking/practices can be introduced
--

3. Checklists like https://deon.drivendata.org/#default-checklist can be useful!
--

4. Keep learning and exploring different perspectives to you own 
---



```{r endslide, child="assets/endslide.Rmd"}
```

