---
title: "ETC5512: Wild Caught Data"
toc: false
format:
  html:
    page-layout: full
---

Data can be obtained from many sources. It may be generated via experiments, collected from observational studies or surveys, obtained via sampling, or recorded using sensors. 

Each type of data has its own characteristics that affect the analysis tools we use. Very large data sets come with their own challenges and require some database skills. 

This unit will equip you with the tools to understand and use different sources of data. Open data sources will be emphasised.

## Learning objectives

1. Understand the definitions, allowed usage, digital identification and licensing of open data

2. Know about common open data sources, how they are used and effectively search for new sources

3. Explain the differences between data collection methods and the limitations for data analysis

4. Work with the range of different data formats of open data, including APIs

5. Understand ethical constraints and privacy limits when working with open data

6. Recognise the components of effective curation needed for open data.

## Teachers

* **Kate Saunders** - Lecturer and Chief Examiner
* **Krisanat Anukarnsakulchularp** - Tutor 
* **Maliny Po** - Tutor


```{r}
#| label: load_packages
#| include: false
#| message: false
#| warning: false
#| echo: false
#| cache: false
library(tidyverse)
options(knitr.kable.NA = '')
source(here::here("course_info.R"))
week <- as.integer(1 + ((Sys.Date() - as_date(start_semester))/7))
```

## Weekly schedule

* Seminar: Monday  2 pm - 4 pm
* Drop in & practice session: Monday 10 - 11 pm (zoom)
* Refer to your timetable for your tutorial time
* Refer to Moodle for room locations and zoom links

## Consults 

* Kate: Monday 11 - 12 pm. 
(Room 64, Level 5, Building H, Caulfield Campus or see Moodle for zoom link)
* Kris: Tuesday  <span style="color:red;"> **2:00 - 3:00 pm** </span> 
(Room 232A Building 6, 29 Ancora Imparo Way Clayton Campus)
* Maliny: Wednesday <span style="color:red;"> **3:30 - 4:30 pm** </span>  
(Room 232A Building 6, 29 Ancora Imparo Way Clayton Campus)
* <span style="color:red;"> **New session** </span> Alternates Kris and Maliny: Thursday 3:30 - 4:30 pm 
(Room 232A Building 6, 29 Ancora Imparo Way Clayton Campus)

## Schedule

```{r}
#| label: schedule2
#| message: false
#| warning: false
#| echo: false
#| output: asis
schedule |>
    transmute(
        Week = if_else(Topic == "Mid-semester break", NA, Week, missing = Week),
        Date = format(Date, "%d %b"),
        Show = file.exists(glue::glue("./week{Week}/index.qmd")),
        Show_Tutorial = file.exists(glue::glue("./week{Week}/tutorial/tutorial-0{Week}.qmd")),
        Show_Solution = file.exists(glue::glue("./week{Week}/tutorial/tutorial-0{Week}-sol.qmd")),
        Topic = if_else(!Show, Topic, glue::glue('[{{Topic}}](./week{{Week}}/index.html){target="_blank"}', .open = "{{", .close = "}}")),
        Tutorial = if_else(!Show_Tutorial, "", glue::glue('[{{< fa list-check >}}](./week{{{Week}}}/tutorial/tutorial-0{{{Week}}}.html){target="_blank"}', .open = "{{{", .close = "}}}")),
        Solution = if_else(!Show_Solution, "", glue::glue('[{{< fa lightbulb >}}](./week{{{Week}}}/tutorial/tutorial-0{{{Week}}}-sol.html){target="_blank"}', .open = "{{{", .close = "}}}"))
        # Reference = glue::glue("[{Reference}]({Reference_URL})")
        # Assessments = if_else(is.na(Assignment), Assignment, glue::glue("[{Assignment}]({File})"))
    ) |>
    select(-Show, -Show_Tutorial, -Show_Solution) |>
    knitr::kable(format = "markdown")
```