---
title: "ETC5512: Wild Caught Data"
subtitle: "Semester 1 2020"
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  message = FALSE,
  warning = FALSE
)
library(tidyverse)
library(knitr)
library(kableExtra)
library(glue)
```

This is the course website, with information about the consultation times, schedule of topics, lecture notes and computer lab instructions.

## Lecturers 

- Dr Didier Nibbering
- Dr Patricia Menéndez
- Dr Emi Tanaka
- Professor Di Cook (Chief Examiner)

## Tutors

- Cynthia Huang
- Stephanie Kobakian
- Asghar Mustafa

## Expectations 

- Unless your question is a personal issue or an administrative matter, please do not send us direct email. The moodle discussion forum is appropriate for questions related to the course.
- For personal or private administrative issues, the email contact is: ETC5512.Clayton-x@monash.edu
- Lectures are recorded, and you are expected to have either attended the lecture, or watched the recordings fully, prior to the computer lab for the week.
- Computer labs are not recorded, and attendance is expected.
- Bring your own laptop to the computer labs. If you **do not** have a laptop you can borrow one from [MoVE](https://www.monash.edu/learning-teaching/innovation/educational-technologies/move)
- **If you register after the start of the semester** or if you
  miss a lecture/lab, it is your responsibility to catch up with
  missed material, learn about due dates for material to be turned in,
  and getting assigned to a group for team work, as necessary.
- The computer software R and RStudio will be used fr the unit.
- Lab quizzes are due each week, prior to the computer lab. Quizzes will be available at flux.qa/REAB3Q. You need to sign in with your Monash student account. If you do not enter your name and Monash student authcate correctly, it will result in n marks for the quiz. Each quiz has five equally-weighted multiple-choice questions. A quiz is due at the start of the computer lab each week. 

## Tentative Schedule

- Lectures: **Wed 2-3pm** and **Wed 4-5pm**, LTB 92. 
- Computer labs: **Thu 2-3pm**, LTB 92. Bring your own computer, with necessary software already installed.

There are no lectures or tutorials during the midsemester break.

```{r}
schedule <- read_csv("materials/schedule.csv") %>% 
  mutate(Slides = case_when(
    !is.na(HTML) & !is.na(PDF) ~ glue("<a class='fa fa-link' href='slides/{HTML}'></a> <a class='fa fa-file-pdf' href='slides/{PDF}'></a>"),
    !is.na(HTML) ~ glue("<a class='fa fa-link' href='slides/{HTML}'></a>"),
    !is.na(PDF) ~ glue("<a class='fa fa-file-pdf' href='slides/{PDF}'></a>"),
    TRUE ~ ""
  ),
  Lab = case_when(
    !is.na(Lab) ~ glue("<a class='fa fa-link' href='{Lab}'></a>"),
    TRUE ~ ""
  ),
  Topic = ifelse(is.na(Topic), "", Topic),
  Materials = ifelse(is.na(Materials), "", Materials)) %>% 
  select(Week, Lecturer, Slides, Lab, Topic, Materials)

midsem_break <- 5 # after which week

schedule %>% 
  mutate(Lecturer = cell_spec(
    Lecturer, "html",
    color = factor(Lecturer, unique(Lecturer), c("#8e4f89", "#4269a5", "#39825a", "#dc6141"))
  )) %>% 
  kable(escape = FALSE, align = c("c", "c", "c", "c", "l", "l")) %>% 
  kable_styling(bootstrap_options = "striped",
                fixed_thead = T) %>% 
  column_spec(1, width = "0.5cm") %>% 
  column_spec(2, width = "1cm") %>% 
  column_spec(3, width = "0.5cm") %>% 
  group_rows("Midsemester Break", midsem_break + 1, 12, label_row_css = "background-color: #d9e1e8;") %>% 
  add_indent(1:midsem_break) # this needs to be done to match up with above
```



