---
title: "ETX2250/ETF5922: An Introduction to Data Visualisation"
toc: false
format:
  html:
    page-layout: full
---

This unit uses data visualisation to explore and analyse data sets of all sizes, and it introduces some business analytic models for interpretation and prediction.

## Learning outcomes

1. Critically evaluate the quality of a data visualisation using the principles of graphical excellence
2. Successfully create high-quality data visualisations for a given dataset using software, such as Power BI or R.
3. Develop the skills to effectively use data visualisations to communicate insights to an audience
4. Develop the skills to effectively use data visualisations to communicate in the appropriate medium, report writing, in presentations and digital storytelling.
5. Apply advanced topics in data visualisation, such as using interactivity and animation to enhance communication.

## Teachers

* **Kate Saunders** - Chief Examiner
* **Krisanat Anukarnsakulchularp** - Instructor
* **Bets Ruscoe** - Instructor
* **Javad Vahdat Atashgah** - Instructor


```{r}
#| label: load_packages
#| include: false
#| message: false
#| warning: false
#| echo: false
#| cache: false
library(tidyverse)
options(knitr.kable.NA = '')
source(here::here("course_info.R"))
week <- as.integer(1 + ((Sys.Date() - as_date(start_semester))/7))
```

## Weekly schedule

* 3 hour workshop, Monday - Wednesday - Friday, 9am-12pm each week

## Consultation

* Monday - Wednesday - Friday, 1pm-3pm each week

```{r}
#| label: schedule2
#| message: false
#| warning: false
#| echo: false
#| output: asis
schedule |>
    transmute(
        Week = if_else(Topic == "Mid-semester break", NA, Week, missing = Week),
        Date = format(Date, "%d %b"),
        Show = file.exists(glue::glue("./week{Week}/index.qmd")),
        Show_Tutorial = file.exists(glue::glue("./week{Week}/tutorial/tutorial-0{Week}.qmd")),
        Show_Solution = file.exists(glue::glue("./week{Week}/tutorial/tutorial-0{Week}-sol.qmd")),
        Topic = if_else(!Show, Topic, glue::glue('[{{Topic}}](./week{{Week}}/index.html){target="_blank"}', .open = "{{", .close = "}}")),
        Tutorial = if_else(!Show_Tutorial, "", glue::glue('[{{< fa list-check >}}](./week{{{Week}}}/tutorial/tutorial-0{{{Week}}}.html){target="_blank"}', .open = "{{{", .close = "}}}")),
        Solution = if_else(!Show_Solution, "", glue::glue('[{{< fa lightbulb >}}](./week{{{Week}}}/tutorial/tutorial-0{{{Week}}}-sol.html){target="_blank"}', .open = "{{{", .close = "}}}"))
        # Reference = glue::glue("[{Reference}]({Reference_URL})")
        # Assessments = if_else(is.na(Assignment), Assignment, glue::glue("[{Assignment}]({File})"))
    ) |>
    select(-Show, -Show_Tutorial, -Show_Solution) |>
    knitr::kable(format = "markdown")
```