---
title: "ETC5512: Wild Caught Data"
week: "Week X"
subtitle: "Subject"
author: "First Last"
email: "xxx@monash.edu"
date: "XXth Mar 2020"
pdflink: ""
bgimg: "images/bg1.jpg"
output:
  xaringan::moon_reader:
    css:
      - ninjutsu 
      - "assets/animate.css"
      - "assets/monash-logo.css"
      - "assets/monash-brand.css"
      - "assets/monash-fonts.css"
      - "assets/styles.css" # small improvements
      - "assets/custom.css" # add your own CSS here!
      - "assets/demo.css" # this should be removed
    self_contained: false 
    seal: false 
    chakra: 'libs/remark-latest.min.js'
    lib_dir: libs
    includes:
      in_header: "assets/custom.html"
    mathjax: "assets/mathjax-local/MathJax.js?config=TeX-AMS-MML_HTMLorMML"
    nature:
      highlightStyle: github 
      highlightLanguage: r 
      highlightLines: true
      highlightSpans: false 
      countIncrementalSlides: false
      slideNumberFormat: '%current%/%total%'
      navigation:
        scroll: false 
        touch: true
        click: false
      ratio: '16:9'
---


```{r titleslide, child="components/titleslide.Rmd"}
```

---

Test fdfsd

---

# Monash EBS Teaching Template

* This is the `xaringan` template for Monash EBS teaching

* It tries to respect the [Monash Brand Guidelines](https://www.monash.edu/brandbook/brand-elements)

---

# Monash Brand Colors

.grid[

.item[

## Primary colors

* <div class="monash-bg-blue inline-box"></div> monash-bg-blue
* <div class="monash-bg-gray80 inline-box"></div> monash-bg-gray80
* <div class="monash-bg-gray50 inline-box"></div> monash-bg-gray50
* <div class="monash-bg-gray10 inline-box"></div> monash-bg-gray10
* <div class="monash-bg-black inline-box"></div> monash-bg-black 
* <div class="monash-bg-white inline-box"></div> monash-bg-white 

Removing the "`bg-`" will change the text color instead.
E.g. .monash-blue[monash-blue].

]
.item[

## Secondary colors

* <div class="monash-bg-blue2 inline-box"></div> monash-bg-blue2
* <div class="monash-bg-purple2 inline-box"></div> monash-bg-purple2
* <div class="monash-bg-fuchsia2 inline-box"></div> monash-bg-fuchsia2
* <div class="monash-bg-ruby2 inline-box"></div> monash-bg-ruby2
* <div class="monash-bg-pink2 inline-box"></div> monash-bg-pink2 
* <div class="monash-bg-red2 inline-box"></div> monash-bg-red2 
* <div class="monash-bg-orange2 inline-box"></div> monash-bg-orange2
* <div class="monash-bg-umber2 inline-box"></div> monash-bg-umber2
* <div class="monash-bg-olive2 inline-box"></div> monash-bg-olive2 
* <div class="monash-bg-green2 inline-box"></div> monash-bg-green2 


]
]


---


```{r endslide, child="components/endslide.Rmd"}
```
