---
pagetitle: "ETC5512"
subtitle: "Case Study: Using Large Language Models for Data Cleaning"
author: Cynthia Huang
email: "ETC5512.Clayton-x@monash.edu"
date: "Wild Caught Data"
department: "Department of Econometrics and Business Statistics"
unit-url: "wcd.numbat.space"
footer: "ETC5512"
format: 
  revealjs: 
    logo: images/monash-one-line-black-rgb.png
    slide-number: c
    multiplex: false
    theme: ../assets/monash.scss
    show-slide-number: all
    show-notes: false
    controls: true
    width: 1280
    height: 720
    css: [../assets/custom.css, ../assets/lecture-01.css]
    include-after-body: "../assets/after-body.html"
    chalkboard:
      boardmarker-width: 5
      buttons: true
editor: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  fig.width = 6,
  fig.height = 4,
  fig.align = "center",
  fig.retina = 3,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  cache = FALSE,
  cache.path = "cache/"
)

```

## <br>[`r rmarkdown::metadata$pagetitle`]{.monash-blue} {#etc5523-title background-image="images/bg-01.png"}

### `r rmarkdown::metadata$subtitle`

Lecturer: *`r rmarkdown::metadata$author`*

`r rmarkdown::metadata$department`

::: tl
<br>

<ul class="fa-ul">

<li><i class="fas fa-envelope"></i>`r rmarkdown::metadata$email`</li>

<li><i class="fas fa-calendar-alt"></i>
`r rmarkdown::metadata$date`</li>

<li><i class="fas fa-globe"></i>
`r rmarkdown::metadata[["unit-url"]]`</li>

</ul>

<br>
:::

## Today's lecture

::: callout-note
## What we'll cover

* The basics inputs and outputs of LLMs. 

* What data cleaning tasks LLMs can assist with. 

* How to appropriately cite and document data and code generated by LLMs. 

:::

. . .

::: callout-note
## Coding Perspective:

* Learn how to generate R code via LLM chat interfaces

* See how to interact with LLMs from within R using the ellmer package

:::

# LLM Basics {background-color="#006DAE"}

## Slide 1 

some stuff

```{r}
#| echo: true
#| eval: true
#| code-line-numbers: "1|2|3|1-3"

x = 1
y = 2
x + y 
```

## Slide 2 

more cool stuff

## Breakout Session

::: {.callout-note style="width: 75%; margin: 0 auto;"}
## Try it yourself time

- Try this thing 

- Also try this thing 

- Feeling adventurous do this

:::

# LLMs for augmenting an analysis {background-color="#006DAE"}

## Slide 1 

some stuff

## Slide 2 

more cool stuff

## Breakout Session

::: {.callout-note style="width: 75%; margin: 0 auto;"}
## Try it yourself time

- Try this thing 

- Also try this thing 

- Feeling adventurous do this

:::

# ellmer {background-color="#006DAE"}

## Slide 1 

some stuff

## Slide 2 

more cool stuff

## Breakout Session

::: {.callout-note style="width: 75%; margin: 0 auto;"}
## Try it yourself time

- Try this thing 

- Also try this thing 

- Feeling adventurous do this

:::

# Wrap Up {background-color="#006DAE"}


## Slide 1 

stuff 

## Slide 2 

stuff 

## Summary 

:::callout-note
## What we've learnt 

:::incremental 

- Learnt the basics of what LLMs can receive as inputs and generate as outputs

- Explored different 'wild caught data' tasks that LLMs can be used for

- Generated LLM outputs for these tasks including code and text data

- Learnt how the package {ellmer} can be used to interact with LLMs from R

- Discussed pitfalls to avoid when using LLMs for data cleaning

:::

:::

# Questions {background-color="#006DAE"}
