---
title: 'ETC5512: Instruction to Open Data'
format: 
  html:
    toc: true
    css: ../../assets/tutorial.css
    embed-resources: true
  pdf:
    toc: true
editor: visual
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Learning Objectives

-   Identify whether data are experimental or observational
-   Delineate the data collection methods
-   Logically suggest the population that a sample represents

### Before your tutorial

**Work** through the following [startR modules](https://startr.numbat.space/):

-   Do the module on Projects and Paths (Module 4). *From this week onward we will assume you know how to use RProjects and why these help us organise our analytics work.*

-   Do the module on Strategies for troubleshooting R (Module 5).

These should take you \~ 50 minutes.

## Package Installation

Ensure you have the packages installed from Week 1's tutorial.

Also install 

```{r}
#| eval: false
install.packages("tibble")
install.packages("maps")
install.packages("ggthemes")
```

## Exercise 1 

<!-- observational or experiment, and according to my taxonomy, describe types of variables, brainstorm questions, given a question define population-->

This question relates to the [Tidy Tuesday Data on locations of alternative fuel recharging stations](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-03-01/readme.md). Have a read through this site, and also visit the link to the data providers, DOT.

### a. About the data

Read the details about the data at [DOT](https://data-usdot.opendata.arcgis.com/datasets/usdot::alternative-fueling-stations/about). How is this data collected, do you think? 

### b. Data type 

What type of data is this? (observational, experimental, survey, census)

### c. Population vs sample

Describe the population, and what is the sample. 

### d. Download and plot the data

Download the data and plot the fueling locations on a map, coloured by fuel type.

```{r}
library(tidyverse)
library(ggthemes)
library(maps)

# Note you can read data directly from a website
stations <- read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-03-01/stations.csv')

# Get the map data for the USA
usa <- map_data("state")

# Filter to continental USA using map boundary
stations <- stations |>
  filter(between(LONGITUDE, min(usa$long)-1, max(usa$long)+1),
         between(LATITUDE, min(usa$lat)-1, max(usa$lat)+1))

# Plot the sites on a map 
# Create a different map for each fuel type 

ggplot() + 
  geom_path(data=usa, aes(x=long, y=lat, group=group), colour="grey80") + 
  geom_point(data=stations, aes(x=LONGITUDE, 
                                y=LATITUDE,
                                colour=FUEL_TYPE_CODE),
             alpha=0.8) +
  facet_wrap(~FUEL_TYPE_CODE, ncol=4) +
  coord_map() +
  theme_map() +
  theme(legend.position = "none")
```

### e. New Fuel Stations

Count the number of new stations by month, and make a time series plot by fuel type. 

```{r}
# Time line of opening
stations |>
  mutate(OPEN_DATE = as.Date(OPEN_DATE)) |>
  filter(!is.na(OPEN_DATE)) |>
  mutate(m = month(OPEN_DATE),
         yr = year(OPEN_DATE)) |>
  mutate(open_yrmth = as.Date(paste(yr, m, "01", sep="-"), "%Y-%M-%d")) |>
  group_by(open_yrmth, FUEL_TYPE_CODE) |>
  summarise(nopen = n(), .groups = "drop") |>
ggplot(aes(x=open_yrmth, 
           y=nopen, 
           colour=FUEL_TYPE_CODE)) +
  geom_line() +
  facet_wrap(~FUEL_TYPE_CODE, ncol=4, scales="free_y") +
  ylab("# opening") +
  scale_x_date("", date_labels="%y") +
  theme(legend.position = "none")
```

### g. Fuel growth

If the question to answer is "which alternative fuel vehicle is the fastest growing?" what is the explanatory (independent, predictor) variable and what is the response variable?

## Exercise 2

Here we will look at the [Chocolate bar ratings](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-01-18/readme.md). Details (brief) of how the data was collected are provided [here](https://github.com/rfordatascience/tidytuesday/issues/408) and more about the data itself is [here](http://flavorsofcacao.com/chocolate_database.html).

```{r}
chocolate <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-01-18/chocolate.csv')
```

### a. Type of data
What type of data is this? (observational, experimental, survey, census)

### b. Data collection
How is the data collected?

### c. Population vs sample 
Describe the population.

### d. Response and predictor variables

For the question "Which country of origin of the bean obtains the best rating?" state the response and predictor variables.

### e. Visualise your data

Make a plot to answer the previous question. (Only use countries with more than 10 records.)

```{r}

# Which countries have more than 10 
keep <- chocolate |>
  count(country_of_bean_origin, sort = TRUE) |>
  filter(n>10) |>
  pull(country_of_bean_origin)

# Filter to those countries
chocolate_countries = chocolate |>
  filter(country_of_bean_origin %in% keep) 

# Plot the results
  ggplot(data = chocolate_countries, aes(x=fct_reorder(country_of_bean_origin, rating, mean), 
             y=rating)) +
    geom_jitter(width=0.1) + 
    stat_summary(fun = mean, fun.min = median, fun.max = median,
                 geom = "point", colour = "orange") +
    xlab("") + 
    coord_flip()
```

## In your own time

Read the description of the study titled ["Clearing the Fog: Is Hydroxychloroquine Effective in Reducing COVID-19 Progression (COVID-19)"](https://clinicaltrials.gov/ct2/show/NCT04491994). 

### a. Data type
What type of data is this? (observational, experimental, survey, census)

### b. Study Participants

How many subjects participated in the study at the start, and to completion?

### c. Study details 

What are the:

- experimental units? 
- factor? 
- blocking factors?
- response variable (outcome measures)?

### d. Study desing

How are subjects assigned to treatment groups?

### e. Study results

What were the results of the study?

### f. Analyse results

Construct the data from the results reported. Compute the proportion of subjects with progression of COVID after 5 days, for the two treatments. Include the standard error of the estimate.

```{r}
hcq <- tibble(trt = c("standard", "standard", "hcq", "hcq"),
              progression = c("all", "yes", "all", "yes"), 
              count = c(151, 5, 349, 11))
hcq |> 
  pivot_wider(names_from = "progression", values_from = "count") |>
  mutate(p = yes/all) |>
  mutate(se = sqrt(p*(1-p)/all))
```

### g. Study conculstions

Based on the proportions and their standard errors, why would the result of the study be that HCQ does NOT improve the outcomes of COVID patients?

### h. Population vs sample

What is the population for this experiment?

##### © Copyright Monash University
