---
title: 'ETC5512: Instructions for Lab 10'
author: "Will Mackey"
date: "Week 10"
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = TRUE, warning = FALSE, message = FALSE)
set.seed(1)
```


**<span style="color:red">Please do [Preparation](#preparation) *before* the tutorial!</span>**

## `r emo::ji("target")` Objectives

In this tutorial, you will learn to

- interrogate your data by studying summary statistics 
- calculate uncertainties of point estimates


## `r emo::ji("wrench")` Preparation  

1. **Read** Sections 13.1-13.4 of [_R for Data Science_](https://r4ds.had.co.nz/relational-data.html)
   
1. **Set up your project folder**

 Prepare a project calling it say `etc5512-week10` with folders `data` and `analysis` in the project root directory. While you don't have to follow the structure strictly, it's a good habit to maintain a project workflow.

1. **Install `tidyverse`**

```{r, eval = FALSE}
install.packages("tidyverse")
```

2. **Check your project structure**

   Your project structure should look like below.
   
```
etc5512-week10
├── README.md
├── analysis
│   └── exercise.Rmd
├── data
└── etc5512-week10.Rproj
```

3. Load the relevant packages

```{r exercise}
library(tidyverse)
library(readxl)
library(lubridate)
```



## `r emo::ji("banknote")` Exercise 10A

**Is there a difference between female and male pay for the same hours in the same occupation?**

* Download detailed income data from the ATO and read into R.

```{r q1}
ato_url <- "https://data.gov.au/data/dataset/23b8c299-a85b-4fc0-a07d-5ed14e23a103/resource/6a989b47-29ce-451a-a9ef-ed57be644b61/download/ts18individual14occupationsextaxableincomerange.xlsx"
```

* For each year, sex and occupation, find the average salary or wage for people who had a salary or wage. 

* Download detailed hours data from the ABS: 

```{r q2}
abs_working_url <- "https://www.abs.gov.au/statistics/labour/employment-and-unemployment/labour-force-australia-detailed/mar-2021/EQ08.xlsx"

```

* For each year, sex and occupation, find the average number of hours worked

* Before joining the datasets by `year`, `sex` and `occupation`: 
explore and reconcile the differences in values in these two datasets.

* Join datasets to find average wage per hour worked by occupation and sex

* Conduct some post-join checks:
  * does the resulting joined dataset have the same number of observations as the ATO dataset used in the join? Why/why not?
  * does the resulting joined dataset have the same number of observations as the ABS dataset used in the join? Why/why not?
  * how could you tell which of the ATO observations weren't matched with an ABS observation? 
  * what number/% of observations weren't matched?
  * what were the ato_join occupations that did not match?
  * why weren't they matched?

* Generate a new variable `wage_rate`, which is average income divided by average wages.

* Explore:
  * average difference in `wage_rate` for females and males in the same occupation. 
  * plot the 20 occupations with the highest and lowest difference in `wage_rate` between females and males.


## `r emo::ji("party")` Exercise 10B

**Better safe than sorry**

* Write a function called `safe_left_join`, which performs as `left_join(a, b)` does,
but checks that the resulting dataframe has the same number of observations as 
dataset `a`. 


