---
title: 'ETC5512: Solutions for Lab 8'
author: "Emi Tanaka"
date: "Week 8"
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = TRUE)
set.seed(1)
filter <- dplyr::filter
```



## `r emo::ji("school")` Exercise 6A

**Interrogate the PISA data**

```{r data, message = FALSE}
library(tidyverse)
library(haven)
pisa2018 <- readRDS("../lectures/data/lecture-06/pisa2018.rds") %>% 
   rename(country = CNT, gender = ST004D01T, birth_year = ST003D03T) %>%
  mutate(country = as_factor(country), gender = as_factor(gender))
```


(a) The number of students who participated by country/region is given below.

```{r count}
pisa2018 %>% 
   group_by(country) %>% 
   tally() %>%  # shortcut for summarise(n = n())
   knitr::kable(col.names = c("Country", "Number of particpating students"))
```

(b) Draw a dot plot with mean differences of math scores on the plausible value `PV1MATH` by gender for each country using the weight `W_FSTUWT`.

```{r plot_dif_ests, fig.height=12}

mathdiff_df <- pisa2018 %>% 
  mutate(gender = as_factor(gender))%>%
  group_by(gender, country) %>% 
  summarise(math = weighted.mean(PV1MATH, 
                                 w = W_FSTUWT)) %>%  #<<
  ungroup() %>% 
  pivot_wider(country, names_from = gender, 
              values_from = math) %>% 
  mutate(diff = Female - Male,
         country = fct_reorder(country, diff)) 

ggplot(mathdiff_df, aes(x = diff, y = country)) + 
  geom_point() + 
  geom_vline(xintercept = 0, color = "red") + 
  labs(y = "Country", 
       x = "Difference in mean PV1 (girl - boy)") 

```

(c) Now calculate this value with each set of replicate weights (e.g., `W_FSTURWT1` - `W_FSTURWT80`)

```{r rep_dif_ests}

rep_mathdiff_df <- pisa2018 %>%
  group_by(gender, country) %>% 
  summarise(dplyr::across(c(W_FSTUWT,W_FSTURWT1:W_FSTURWT80), ~ weighted.mean(PV1MATH, w = .x ))) %>% 
  ungroup() %>% 
  pivot_longer(cols = W_FSTURWT1:W_FSTURWT80, 
               names_to = "replicate_ests",
               values_to = "PV1MATH") %>%
  pivot_wider(id_cols = c(country,replicate_ests,W_FSTUWT),
              names_from = gender, 
              values_from = c(PV1MATH, W_FSTUWT)) %>% 
    mutate(rep_female_male_diff = PV1MATH_Female - PV1MATH_Male,
           wtd_female_male_diff = W_FSTUWT_Female - W_FSTUWT_Male)

```

(d) Use the formula provided on the slide `Using replicate weights with the PISA` to calculate the variance around this estimate. 

```{r use_rep_wts}
final_df_math_dif <- rep_mathdiff_df %>%  
  group_by(country,wtd_female_male_diff) %>%
  summarise(math_dif_var = 1/20*sum((rep_female_male_diff - wtd_female_male_diff)^2)) %>%
  ungroup()

```

Note that we are just considering PV1MATH (rather than math performance overall as in lectures) so we don't need to consider measurement (unlike the lecture).

(e)  Add error bars to your plot form (b) based on 95% confidence intervals. 

```{r error_plot_math, fig.height = 12}
final_df_math_dif %>%  
  mutate(country = fct_reorder(country, wtd_female_male_diff)) %>% 
  ggplot(aes(x = wtd_female_male_diff, y = country)) + 
  geom_point() + 
  geom_errorbar(aes(xmin = wtd_female_male_diff - 1.96*sqrt(math_dif_var),  #<<
                    xmax = wtd_female_male_diff + 1.96*sqrt(math_dif_var))) +
  geom_vline(xintercept = 0, color = "red") + 
  labs(y = "Country", 
       x = "Difference in mean PV1 (girl - boy)") 

```

### What about students who only answered a math component?

(f) How many students conducted the test with `BOOKID` as Form 1-12 or 67-78 (i.e. the number of students who sat for a test that included a maths component) by each country?

```{r count-maths}
pisa2018 %>% 
   filter(BOOKID %in% paste("Form", c(1:12, 67:78))) %>% 
   group_by(country) %>% 
   tally() %>% 
   knitr::kable(col.names = c("Country", "Number of students who undertook the maths component"))
```

(g) Draw the same plot as b. but just based on math scores for students whose test did include a maths component.

```{r error_plot_math2, fig.height = 12, cache = TRUE}

 pisa2018 %>% 
  mutate(BOOKID = as_factor(BOOKID)) %>%
  filter(BOOKID %in% paste("Form", c(1:12, 67:78))) %>% 
  mutate(gender = as_factor(gender))%>%
  group_by(gender, country) %>% 
  summarise(dplyr::across(c(W_FSTUWT,W_FSTURWT1:W_FSTURWT80), ~ weighted.mean(PV1MATH, w = .x ))) %>%  
  ungroup() %>% 
  pivot_longer(cols = W_FSTURWT1:W_FSTURWT80, 
               names_to = "replicate_ests",
               values_to = "PV1MATH") %>%
  pivot_wider(id_cols = c(country,replicate_ests,W_FSTUWT),
              names_from = gender, 
              values_from = c(PV1MATH, W_FSTUWT)) %>% 
  mutate(rep_female_male_diff = PV1MATH_Female - PV1MATH_Male,
           wtd_female_male_diff = W_FSTUWT_Female - W_FSTUWT_Male)%>%  
  group_by(country,wtd_female_male_diff) %>%
  summarise(math_dif_var = 1/20*sum((rep_female_male_diff - wtd_female_male_diff)^2)) %>%
  ungroup()%>%  
  mutate(country = fct_reorder(country, wtd_female_male_diff)) %>% 
  ggplot(aes(x = wtd_female_male_diff, y = country)) + 
  geom_point() + 
  geom_errorbar(aes(xmin = wtd_female_male_diff - 1.96*sqrt(math_dif_var),  #<<
                    xmax = wtd_female_male_diff + 1.96*sqrt(math_dif_var))) +
  geom_vline(xintercept = 0, color = "red") + 
   labs(title = "d. Mean math score gender gap by country (for students who took the maths test)",
        y = "", x = "Mean difference between genders (Girl - Boy) in math score") 
```

## `r emo::ji("student")` Exercise 6B

Using `PV1READ`

```{r error_plot_read, fig.height=12}
 pisa2018 %>% 
  mutate(BOOKID = as_factor(BOOKID)) %>%
  mutate(gender = as_factor(gender))%>%
  group_by(gender, country) %>% 
  summarise(dplyr::across(c(W_FSTUWT,W_FSTURWT1:W_FSTURWT80), ~ weighted.mean(PV1READ, w = .x ))) %>%  
  ungroup() %>% 
  pivot_longer(cols = W_FSTURWT1:W_FSTURWT80, 
               names_to = "replicate_ests",
               values_to = "PV1READ") %>%
  pivot_wider(id_cols = c(country,replicate_ests,W_FSTUWT),
              names_from = gender, 
              values_from = c(PV1READ, W_FSTUWT)) %>% 
  mutate(rep_female_male_diff = PV1READ_Female - PV1READ_Male,
           wtd_female_male_diff = W_FSTUWT_Female - W_FSTUWT_Male)%>%  
  group_by(country,wtd_female_male_diff) %>%
  summarise(math_dif_var = 1/20*sum((rep_female_male_diff - wtd_female_male_diff)^2)) %>%
  ungroup()%>%  
  mutate(country = fct_reorder(country, wtd_female_male_diff)) %>% 
  ggplot(aes(x = wtd_female_male_diff, y = country)) + 
  geom_point() + 
  geom_errorbar(aes(xmin = wtd_female_male_diff - 1.96*sqrt(math_dif_var),  #<<
                    xmax = wtd_female_male_diff + 1.96*sqrt(math_dif_var))) +
  geom_vline(xintercept = 0, color = "red") + 
   labs(title = "d. Mean reading score gender gap by country",
        y = "", x = "Mean difference between genders (Girl - Boy) in reading score") 
```

Using `PV1SCIE`

```{r error_plot_science, fig.height = 12, cache = TRUE}

 pisa2018 %>% 
  mutate(BOOKID = as_factor(BOOKID)) %>%
  filter(BOOKID %in% paste("Form", c(1:12, 67:78))) %>% 
  mutate(gender = as_factor(gender))%>%
  group_by(gender, country) %>% 
  summarise(dplyr::across(c(W_FSTUWT,W_FSTURWT1:W_FSTURWT80), ~ weighted.mean(PV1SCIE, w = .x ))) %>%  
  ungroup() %>% 
  pivot_longer(cols = W_FSTURWT1:W_FSTURWT80, 
               names_to = "replicate_ests",
               values_to = "PV1SCIE") %>%
  pivot_wider(id_cols = c(country,replicate_ests,W_FSTUWT),
              names_from = gender, 
              values_from = c(PV1SCIE, W_FSTUWT)) %>% 
  mutate(rep_female_male_diff = PV1SCIE_Female - PV1SCIE_Male,
           wtd_female_male_diff = W_FSTUWT_Female - W_FSTUWT_Male)%>%  
  group_by(country,wtd_female_male_diff) %>%
  summarise(math_dif_var = 1/20*sum((rep_female_male_diff - wtd_female_male_diff)^2)) %>%
  ungroup()%>%  
  mutate(country = fct_reorder(country, wtd_female_male_diff)) %>% 
  ggplot(aes(x = wtd_female_male_diff, y = country)) + 
  geom_point() + 
  geom_errorbar(aes(xmin = wtd_female_male_diff - 1.96*sqrt(math_dif_var),  #<<
                    xmax = wtd_female_male_diff + 1.96*sqrt(math_dif_var))) +
  geom_vline(xintercept = 0, color = "red") + 
   labs(title = "d. Mean science score gender gap by country",
        y = "", x = "Mean difference between genders (Girl - Boy) in science score") 
```

From the plots we can see that girls in general perform better than boys in reading and science!

## `r emo::ji("teacher")` Exercise 6C

```{r better-plot, fig.height = 12, warning = FALSE, cache = T}
ncountry <- length(unique(pisa2018$country))

math_df <- pisa2018 %>% 
  mutate(BOOKID = as_factor(BOOKID)) %>%
  filter(BOOKID %in% paste("Form", c(1:12, 67:78))) %>% 
  mutate(gender = as_factor(gender))%>%
  group_by(gender, country) %>% 
  summarise(dplyr::across(c(W_FSTUWT,W_FSTURWT1:W_FSTURWT80), ~ weighted.mean(PV1MATH, w = .x ))) %>%  
  ungroup() %>% 
  pivot_longer(cols = W_FSTURWT1:W_FSTURWT80, 
               names_to = "replicate_ests",
               values_to = "PV1MATH") %>%
  pivot_wider(id_cols = c(country,replicate_ests,W_FSTUWT),
              names_from = gender, 
              values_from = c(PV1MATH, W_FSTUWT)) %>% 
  mutate(rep_female_male_diff = PV1MATH_Female - PV1MATH_Male,
           wtd_female_male_diff = W_FSTUWT_Female - W_FSTUWT_Male)%>%  
  group_by(country,wtd_female_male_diff) %>%
  summarise(math_dif_var = 1/20*sum((rep_female_male_diff - wtd_female_male_diff)^2)) %>%
  ungroup()

math_df %>% 
   mutate(country = fct_reorder(country, wtd_female_male_diff),
          avg = wtd_female_male_diff,
          lower = wtd_female_male_diff - 1.96*sqrt(math_dif_var),
          upper = wtd_female_male_diff + 1.96*sqrt(math_dif_var),
          color = case_when(
          avg > 0 & lower > 0 ~ "purple",
          avg < 0 & upper < 0 ~ "blue",
          TRUE ~ "black"
   )) %>% 
   ggplot(aes(avg, country)) + 
   geom_errorbar(aes(xmin = lower, 
                     xmax = upper,
                     color = I(color))) + 
   geom_point() + 
   geom_vline(xintercept = 0, color = "red") + 
   labs(x = "The difference in mean math scores by gender",
        y = "") + 
   scale_x_continuous(breaks = seq(-20, 20, by = 5),
                      labels = c(seq(20, 0, by = -5), seq(5, 20, by = 5))) + 
   annotate("text", label = c("Boys higher", "Girls higher"), 
            x = c(-10, 10), y = -2) + 
   ggtitle("Gender gap in math scores by country") +
   coord_cartesian(clip = "off", ylim = c(1, ncountry - 1)) + 
   theme(plot.margin = unit(c(1,1,3,1), "lines"),
         axis.title.x = element_text(size = 14, face = "bold", margin = unit(c(3, 0, 0, 0), "lines")),
         plot.title = element_text(size = 18, face = "bold"))
```



