---
title: 'ETC5512: Instructions for Tutorial 8'
author: "Emi Tanaka"
date: "Week 8 Australian election data"
output: 
  html_document:
    css: tutorial.css
---

## `r emo::ji("target")` Objectives

In this tutorial, you will learn to

- work with map data and visualise it
- integrate information from another source


## `r emo::ji("wrench")` Preparation  

Note: you should do Preparation **_before_** the tutorial.

1. **Installing relevant R-package**

```{r, eval = FALSE}
install.packages("sf")
```
2. **Get the distribution of preferences by candidate by division for the 2019 Australian Federal Election**
   * Go to https://results.aec.gov.au
     * 2019 federal election
     * Downloads
     * Distribution of preferences by candidate by division
   * Or directly here: https://results.aec.gov.au/24310/Website/Downloads/HouseDopByDivisionDownload-24310.csv


3. **Getting the electoral district geographical boundaries**
    * Go to https://www.aec.gov.au/electorates/gis/licence.htm to agree with the license
    * Download the ESRI (.shp) zip file for Victoria.
4. **Getting the SA1 geographical boundaries**
   * Go to https://datapacks.censusdata.abs.gov.au/geopackages/ selecting the following options:
      * Employment, Income and Unpaid Work (EIUW)
      * EIUW GeoPackage A



## `r emo::ji("fencing")` Exercise 8A

**Map the election data**

We will draw a map of Victoria and colour the electorate districts with the political party that won that district in the 2019 federal election.

* First, import the election data and map data as below.

```{r, message = FALSE, warning = FALSE}
library(tidyverse)
library(sf)
```


```{r, message = FALSE, warning = FALSE}
election2019 <- read_csv("https://results.aec.gov.au/24310/Website/Downloads/HouseDopByDivisionDownload-24310.csv", 
                             skip = 1) %>% 
   # select winning candidate only 
   filter(CalculationType == "Preference Count" & 
             Elected == "Y" & 
             CountNumber == 0) %>% 
   # make division name upper case so 
   # it can be joined later with map data
   mutate(DivisionNm = toupper(DivisionNm))
```

```{r}
vic_map <- read_sf(here::here("data/vic-july-2018-esri/E_AUGFN3_region.shp")) %>% 
   # to match up with election data
   mutate(DivisionNm = toupper(Elect_div)) 
```



```{r}
# define the political colors (feel free to change)
party_colors <- c("ALP" = "#DE3533", "LNP" = "#ADD8E6", 
                  "KAP" = "#8B0000", "GRN" = "#10C25B", 
                  "GVIC" = "#10C25B", "XEN" = "#ff6300", 
                  "LP" = "#0047AB", "NP" = "#0a9cca", 
                  "IND" = "#000000")
```


```{r vicmap19, eval = F, echo = F}
ggplot(left_join(vic_map, election2019, by = "DivisionNm")) +
   geom_sf(aes(geometry = geometry, fill = PartyAb), 
                  color = "white") + 
   coord_sf(xlim = c(144.8, 145.2), ylim = c(-38.1, -37.6)) +
   scale_fill_manual(values = party_colors) + 
   ggtitle("Winners of Australian Federal Election in 2019", 
           subtitle = "Victoria")
```

* Draw a map that looks like below. The map shows the coordinates with longitude of 144.8 to 145.2 and latitude of -38.1 to -37.6.


```{r vicmap19, eval = T, echo = F}
```

<details style = "padding-left:50px;">
<summary>Solution</summary>

```{r vicmap19, eval = F, echo = T}
```

</details>



## `r emo::ji("snowboarder")` Exercise 8B


**Integrate data from different sources**


* Import the GeoPackage data with SA1 regions.

```{r}
SA1map <- read_sf(here::here("data/Geopackage_2016_EIUWA_for_VIC/census2016_eiuwa_vic_short.gpkg"), 
                  layer = "census2016_eiuwa_vic_sa1_short")
```


* Calculate the centroids for each SA1 region by completing the code below. What is each function doing below?

```{r, eval = FALSE}
SA1map <- SA1map %>% 
   mutate(centroid = map(..., ~setNames(st_centroid(.x), c("x", "y")))) %>%
               unnest_wider(...)
```


* Plot the median weekly personal income data on top of the electoral boundaries like below. Can you find which regions have wealthy individuals? 

```{r income-map, echo = FALSE, message = FALSE, warning = FALSE}
SA1map <- SA1map %>% 
   mutate(centroid = map(geom, ~setNames(st_centroid(.x), c("x", "y")))) %>%
               unnest_wider(centroid)

ggplot(SA1map) +
   geom_point(aes(x, y, color = Median_tot_prsnl_inc_weekly), shape = 3) + 
   geom_sf(data = vic_map, aes(geometry = geometry), fill = "transparent", size = 1.3, color = "black") +
   coord_sf(xlim = c(144.8, 145.2), ylim = c(-38.1, -37.6)) +
  scale_color_viridis_c(name = "Median weekly\nperson income", option = "magma")
```

* Looking at the map in 8A, there is one electorate won by the Green party. Looking at the map above, this is where a lot of wealthy individuals live. Can you say that those who vote for the Green party are rich individuals? Why or why not? Discuss with your classmate. 
* Why do some geographical regions look sparse in terms of the census reported median weekly personal income?
* What is **ecological fallacy**? How does it relate to your conclusion from before?
