---
title: 'ETC5512: Solutions for Tutorial 4'
author: "Di Cook"
date: "Week 4 Airline case study"
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  eval = TRUE,
  message = FALSE,
  warning = FALSE, 
  error = FALSE,
  fig.width = 8,
  fig.height = 6,
  out.width = "100%",
  fig.retina = 3
  )
```


```{r}
# Code needs to run to use in  evaluated chunks
library(tidyverse)
library(DBI)
library(RSQLite)
library(ggmap)
library(ggthemes)
library(gridExtra)
library(forcats)
# Time zone matching
library(lutz)
library(lubridate)
```

## `r emo::ji("minidisc")` Exercise 4A

Build an sqlite database from your month of data.

a. Read the data into R. It's possible to do this, because this is a small amount of data.
b. Create an sqlite data base called "flights" from the data
c. Do a simple query, collecting all the records for the first day of the month
    
```{r eval=FALSE, echo=TRUE}
# This code only needs to run it once
# Read ORIGINAL data - your data might have a different filename
d <- read_csv(here::here("data/994615142_T_ONTIME_REPORTING Jan 2020.csv"))
# Create an SQLite database
con <- dbConnect(RSQLite::SQLite(), "data/:flights:")
copy_to(con, d, "flights", 
  temporary = FALSE, 
  indexes = list(
    c("FL_DATE", 
    "OP_CARRIER", 
    "TAIL_NUM",
    "ORIGIN",
    "DEST"
  )
))

# Check that database is set up
flights_db <- tbl(con, "flights")
jan1 <- flights_db  %>%
  filter(DAY_OF_MONTH==1) %>% 
  select(DAY_OF_MONTH, ORIGIN, DEST, 
         DEP_TIME, ARR_TIME,
         OP_CARRIER, TAIL_NUM) %>%
  collect()
jan1
```


## `r emo::ji("abacus")` Exercise 4B

Explore your data! These tasks should be done using the dplyr interface, so that the **tidy wrangling verbs** can be used instead of raw SQL functions.

a. Find the carrier that had the most flights during the month. Determine which carrier this is? 

```{r}
con <- dbConnect(RSQLite::SQLite(), "data/:flights:")
carrier <- tbl(con, "flights") %>%
  group_by(OP_CARRIER) %>%
  tally(sort=TRUE) %>%
  collect()
carrier %>% head(5)
```

**WN = Southwest had the most flights.**

b. Which airport had the most departing traffic?

```{r}
origin <- tbl(con, "flights") %>%
  group_by(ORIGIN) %>%
  tally(sort=TRUE) %>%
  collect()
origin %>% head(5)
```

**LAX = Los Angeles had the most departing flights.**

c. Compute the smallest, largest and median departure delay for the busiest airport. What would it mean if the median departure delay was negative?

```{r}
lax <- tbl(con, "flights") %>%
  filter(ORIGIN == "LAX") %>%
  collect()
lax_delays <- lax %>% 
  summarise(m1 = min(DEP_DELAY, na.rm=TRUE), 
                 m2 = max(DEP_DELAY, na.rm=TRUE), 
                 m3 = median(DEP_DELAY, na.rm=TRUE))
lax_delays
```

**LAX is the busiest airport. The smallest departure delay for the month was** `r lax_delays$m1` **minutes, which means the flight left early, quite early. The longest delay was** `r lax_delays$m2` **minutes, a day delay. (This looks like a time zone calculation error.) The median delay was** `r lax_delays$m3`**, which is less than 0. That means that 50% of the flights left before scheduled.**

d. Make a side-by-side boxplot or violin plot of the delays for each carrier, at the busiest airport. 
    i. Think about transforming delay because it has a skewed distribution. (If you use a transformation on the axis, check the number of missings. It may be that a lot of data is excluded and you need to do the transformation with `mutate`.)
    ii. Sort the carrier axis by the median delay (this is tricky! Hint: use the `forcats` package).
    iii. Make nice labels on the axis
    iv. Write a paragraph on what is learned about the delays by carrier
    
```{r}
lax %>%
  mutate(DEP_DELAY_log = log10(DEP_DELAY+30)) %>%
  mutate(OP_CARRIER = fct_reorder(as_factor(OP_CARRIER), DEP_DELAY_log, 
                median, na.rm=TRUE)) %>%
  ggplot(aes(x=OP_CARRIER, y=DEP_DELAY_log)) + 
  geom_point(alpha=0.3) + 
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75), fill = "orange") + 
  xlab("carrier") + 
  ylab("Departure delay (log)") +
  theme_bw()
```

**Overall, there is not much difference in the median delays, and the variation in delays between carriers. American Airlines (AA) and Southwest (WN) have the highest median delay and Frontier Airlines (F9) has the lowest median delay. Skywest (OO) has the flight with the longest delay.** 

e. How many records, of the busiest airport, have missing values for departure delay?

```{r}
lax %>% count(is.na(DEP_DELAY)) 
```


## `r emo::ji("round_pushpin")` Exercise 4C

Here we are going to add a new table with airport information, and use this to make a map of flights.

a. Read the airport location data into R, and add a table to your database. 

```{r eval=FALSE, echo=TRUE}
# Only need to do this once
airports <- read_csv(here::here("data","402312038_T_MASTER_CORD.csv")) %>%
  select(-X29)
copy_to(con, airports, "airports", 
  temporary = FALSE
)
```

b. Plot the locations on a map. You should filter the airports to only the latest location. Airports sometimes move `r emo::ji("hand_over_mouth")`. An Open Street Map can be downloaded using the `get_map()` function in the `ggmap` package.

```{r}
dbListTables(con)
airport_locations <- tbl(con, "airports") %>%
  filter(AIRPORT_IS_LATEST == 1, AIRPORT_IS_CLOSED != 1,
         AIRPORT_COUNTRY_CODE_ISO == "US") %>%
  select(AIRPORT, DISPLAY_AIRPORT_NAME, LONGITUDE, LATITUDE) %>%
  collect() #<<
usa_bbox <- c(-130, # min long
              20, # min lat
              -60, # max long
              55) # max lat
usa_map <- get_map(location = usa_bbox, source = "osm")
# Only continental locations will be considered
airport_locations <- airport_locations %>%
  filter(LONGITUDE>usa_bbox[1], LONGITUDE<usa_bbox[3], 
         LATITUDE>usa_bbox[2], LATITUDE<usa_bbox[4])
ggmap(usa_map) + 
  geom_point(data=airport_locations, aes(x=LONGITUDE, LATITUDE),
             colour="#746FB2", alpha=0.5, size=2) +
  theme_map()
```

c. Now the fun part, lets take a day's worth of flights, and plot all the flights. You will need to join the day of flights data with the airport locations, using both the ORIGIN and DESTination.

```{r}
flights_db <- tbl(con, "flights")
jan1 <- flights_db  %>%
  filter(DAY_OF_MONTH==1) %>% 
  select(DAY_OF_MONTH, ORIGIN, DEST, DEP_TIME, ARR_TIME,
         OP_CARRIER, TAIL_NUM) %>%
  collect()
# Remove flights that have a missing ORIGIN or DEST lat/long
# because they are outside continental US
jan1_flights <- jan1 %>% 
  left_join(airport_locations, by=c("ORIGIN" = "AIRPORT")) %>% 
  rename(ORIGIN_lon = LONGITUDE, ORIGIN_lat = LATITUDE, 
         ORIGIN_name = DISPLAY_AIRPORT_NAME) %>%
  left_join(airport_locations, by=c("DEST" = "AIRPORT")) %>%
  rename(DEST_lon = LONGITUDE, DEST_lat = LATITUDE, 
         DEST_name = DISPLAY_AIRPORT_NAME) %>%
  filter(!is.na(ORIGIN_lon)) %>%
  filter(!is.na(DEST_lon)) %>%
  filter(!is.na(DEP_TIME))
ggmap(usa_map) + geom_segment(data=jan1_flights, 
                     aes(x=ORIGIN_lon, 
                         xend=DEST_lon,
                         y=ORIGIN_lat,
                         yend=DEST_lat), 
                     colour="#9651A0", alpha=0.01) +
  geom_point(data=jan1_flights, aes(x=ORIGIN_lon, ORIGIN_lat),
             colour="#746FB2", alpha=0.1, size=1) +
  theme_map()
```

d. Choose the two major carriers for your day of data, and make two separate maps of flights, one for each carrier. Compare and contrast the carrier flight patterns. 

```{r fig.height=4}
p1 <- ggmap(usa_map) + 
  geom_segment(data=filter(jan1_flights, 
                           OP_CARRIER == "DL"), 
                     aes(x=ORIGIN_lon, 
                         xend=DEST_lon,
                         y=ORIGIN_lat,
                         yend=DEST_lat), 
                     colour="#9651A0", alpha=0.1) +
  geom_point(data=jan1_flights, 
             aes(x=ORIGIN_lon, ORIGIN_lat),
             colour="#746FB2", alpha=0.1, size=1) +
  theme_map() + ggtitle("Delta")
p2 <- ggmap(usa_map) + 
  geom_segment(data=filter(jan1_flights, 
                           OP_CARRIER == "WN"), 
                     aes(x=ORIGIN_lon, 
                         xend=DEST_lon,
                         y=ORIGIN_lat,
                         yend=DEST_lat), 
                     colour="#9651A0", alpha=0.1) +
  geom_point(data=jan1_flights, 
             aes(x=ORIGIN_lon, ORIGIN_lat),
             colour="#746FB2", alpha=0.1, size=1) +
  theme_map() + ggtitle("Southwest")
grid.arrange(p1, p2, ncol=2)
```

**I chose Delta and Southwest. It looks a little like Delta has more of a hub system. We see this because both airlines are high volume carriers, but spatially it looks like Southwest has more flights than Delta. Delta flights are operating between fewer airports, and Southwest is more distributed, serving many more airports.**

e. ADVANCED: Now we are going to examine change in patterns over the course of a day. You will need to convert departure time into a standard time. Then break it into  one of four categories: midnight-6am, 6am-noon, noon-6pm, 6pm-midnight. Using all the carriers again, make separate maps for each quarter of the day.  Compare the traffic over these four time blocks.

```{r}
jan1_flights <- jan1_flights %>% 
  mutate(ORIGIN_tz = tz_lookup_coords(ORIGIN_lat, 
                              ORIGIN_lon, warn=FALSE),
         DEST_tz = tz_lookup_coords(DEST_lat, 
                              DEST_lon, warn=FALSE)) %>%
  mutate(DateTime = paste0("2020-02-01 ", 
                           substr(DEP_TIME, 1, 2),":", 
                           substr(DEP_TIME, 3, 4), ":00"))
# Can't keep datetime with different time zones in the same variable
jan1_flights <- jan1_flights %>%
  mutate(
    DateTime_NY = force_tzs(ymd_hms(DateTime),
      tzones = ORIGIN_tz, tzone_out = "America/New_York"))

# Now break into four time categories
jan1_flights <- jan1_flights %>% 
  mutate(TimeOfDay = case_when(
    hour(DateTime_NY) < 7 ~ "early",
    between(hour(DateTime_NY), 7, 11) ~ "morning",
    between(hour(DateTime_NY), 12, 17) ~ "afternoon",
    hour(DateTime_NY) > 17 ~ "evening")
  ) %>%
  mutate(TimeOfDay = factor(TimeOfDay, levels=c("early", "morning", "afternoon", "evening")))
ggmap(usa_map) + 
  geom_point(data=jan1_flights, aes(x=ORIGIN_lon, y=ORIGIN_lat),
             colour="#746FB2", alpha=0.1, size=2) +
  facet_wrap(~TimeOfDay, ncol=2) +
  theme_map()

```

**There's not a lot to see in four big groups like this. Its an exercise in working with time. And also in ordering the four groups appropriately.**

f. ADVANCED: Use the standardised times to follow the path of one plane during the day.


```{r}
# Track the path of one plane
N243WN <- jan1_flights %>% 
  filter(TAIL_NUM == "N243WN") %>%
  arrange(DateTime_NY)
ggmap(usa_map) + 
  geom_segment(data=N243WN, 
               aes(x=ORIGIN_lon, y=ORIGIN_lat, 
                   xend=DEST_lon, yend=DEST_lat),
               colour = "orange", size=2) +
  theme_map()
```

**Plane N243WN have 8 flights during the day, but it goes back and forth between three airports.**
