---
title: 'ETC5512: Instructions for Tutorial 7'
author: "Emi Tanaka"
date: "Week 7 Australian Census"
output: 
  html_document:
    toc: true
---


```{r, include = FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, eval = FALSE)
```


## `r emo::ji("target")` Objectives

In this tutorial, you will learn 

- a bit about how to structure your projects
- to wrangle messy data into tidy data


## `r emo::ji("wrench")` Preparation  


1. **Set up your project folder**

   * If you haven't already, you should have an R project already to work through your exercise. Below is an example of how to get started. 

   * RStudio > New Project > New Directory > New Project
   
     Save it in a directory of your choosing. 
   
     You can call the project what you like, e.g. `etc5512`.
   
   * Create folders called `data` and `analysis` (or similarly named) in the project root directory.

   
   * Create an `exercise-07.Rmd` under the `analysis` (or similarly named) folder.



1. **Getting the census data**
  
   * Go to https://datapacks.censusdata.abs.gov.au/datapacks/ selecting the following options:
     * 2016 Census Datapacks
     * General Community Profile 
     * All geographies
     * Vic 

     The downloaded file would be called `2016_GCP_ALL_for_Vic_short-header.zip` with size of 118 MB.
   
   * Unzip this file and save the folder in under `data` folder.
    

1. **Installing relevant R-packages**

```{r, eval = FALSE}
install.packages(c("remotes", "usethis", "here", "skimr", "unglue"))
```

2. **Check your project structure**

   Your project structure should look like below.

```
etc5512
├── analysis
│   └── exercise-07.Rmd
├── data
│   └── 2016_GCP_ALL_for_Vic_short-header
│       └── 2016\ Census\ GCP\ All\ Geographies\ for\ VIC
│           ├── SA1
│           │   └── VIC
│           │       ├── 2016Census_G17A_VIC_SA1.csv
│           │       ├── 2016Census_G17B_VIC_SA1.csv
│           │       └── 2016Census_G17C_VIC_SA1.csv
│           └── STE
│               └── VIC
│                   ├── 2016Census_G17A_VIC_STE.csv
│                   ├── 2016Census_G17B_VIC_STE.csv
│                   └── 2016Census_G17C_VIC_STE.csv
└── etc5512.Rproj
```

   * Note that it's not desirable to have spaces in your file or folder name, as you may need escape spaces by preceding with `\` like above
   * <span style="color:red">Put your analysis from the exercises into the `analysis/exercise.Rmd`.</span>
   
## `r emo::ji("lifter")` Exercise 7A

**Practice regular expression**

Consider the following vector `cats`. The first letter has information on the sex, followed by minimum income, maximum income, minimum age, and maximum age separated by `_`. Wrangle this vector to extract the information on its own vector or a column in a data frame.

```{r}
cats <- c("F_300_399_15_19_yrs", "F_1_149_35_44_yrs", "M_2000_2999_75_84_yrs" )
```




## `r emo::ji("lifter")` Exercise 7B

**Tidy your census data**

Start with loading the `tidyverse` package and setting hte path ways to the data. 

```{r setup, eval = TRUE}
library(tidyverse)
library(unglue)
filter <- dplyr::filter
census_path <- here::here("data/2016_GCP_ALL_for_Vic_short-header/2016 Census GCP All Geographies for VIC/{geo}/VIC/")
STE_paths <- glue::glue(census_path, "2016Census_G17{alpha}_VIC_STE.csv", 
                        geo = "STE", alpha = c("A", "B", "C"))
SA1_paths <- glue::glue(census_path, "2016Census_G17{alpha}_VIC_SA1.csv", 
                        geo = "SA1", alpha = c("A", "B", "C"))
```

To load the data we will use the `here` function from the `here` package `r emo::ji("package")`. Note that we can use a function from a package without loading the package (i.e. without `library(package)`) by using the form `package::function`. The path argument parsed to `here::here` will be referenced with respect to `Rproj` is. This is particularly helpful if your project structure becomes deeply nested. 

```
etc5512-week07
├── analysis
│   └── exercise-07.Rmd      # << analysis are here
├── data                  # << data are here 
│               
│  
└── etc5512.Rproj  # << make reference point `here`
```


First focus on `2016Census_G17A_VIC_STE.csv` alone. 

To examine the data in the first instance use `skimr::skim`.

```{r}
STE_G17A <- read_csv(STE_paths[1])
skimr::skim(STE_G17A)
```
* How many rows and columns does `STE_G17A` contain?
* Remove any columns with the header that contains the strings `Tot` (totals) or `PI_NS` (personal income not stated).
* Wrangle your data so that it has the columns: `count`, `sex`, `age_min`, `age_max` `income_min`, and `income_max`. Below in the hint is some code to get you started (you will need to replace all the `...`s). For those with no upper bound (e.g. 85 over, 3000 or more), you can use `Inf` in R to signify $\infty$.

<details style = "padding-left:50px;">
<summary>Hint</summary>
```{r, eval = FALSE}
STE_G17Atidy <- pivot_longer(STE_G17A, -STE_CODE_2016, 
                             names_to = "category", values_to = "count") %>% 
   filter(!str_detect(category, "Tot") | ... ) %>% 
   unglue_unnest(category, 
                   c("{sex=[MFP]}_{income_min=\\d+}_{income_max=(\\d+|more)}_{age_min=\\d+}_{age_max=\\d+}_yrs",
                     "{sex=[MFP]}_{income_min=\\d+}_{income_max=(\\d+|more)}_{age_min=\\d+}ov",
                     "{sex=[MFP]}_{income_min=(Neg_Nil_income)}_{age_min=\\d+}_{age_max=\\d+}_yrs",
                     "{sex=[MFP]}_{income_min=(Negtve_Nil_incme)}_{age_min=\\d+}_yrs_ovr"), 
                   remove = FALSE) %>% 
   mutate(...)
```
</details>


The first 10 rows of `STE_G17Atidy` should look similar to below.


```{r, echo = FALSE, eval = TRUE, message = FALSE}
clean_G17 <- function(df, id) {
   df %>% 
      pivot_longer(-{{id}},
                   names_to = "category",
                   values_to = "count") %>% 
      filter(!str_detect(category, "Tot") & !str_detect(category, "PI_NS")) %>%
      unglue_unnest(category, 
                   c("{sex=[MFP]}_{income_min=\\d+}_{income_max=(\\d+|more)}_{age_min=\\d+}_{age_max=\\d+}_yrs",
                     "{sex=[MFP]}_{income_min=\\d+}_{income_max=(\\d+|more)}_{age_min=\\d+}ov",
                     "{sex=[MFP]}_{income_min=(Neg_Nil_income)}_{age_min=\\d+}_{age_max=\\d+}_yrs",
                     "{sex=[MFP]}_{income_min=(Negtve_Nil_incme)}_{age_min=\\d+}_yrs_ovr"), 
                   remove = FALSE) %>% 
      mutate(income_max = as.numeric(income_max),
             income_max = case_when(str_detect(income_min, "Neg") ~ 0, 
                                               is.na(income_max) ~ Inf,
                                                            TRUE ~ income_max),
             income_min = ifelse(income_max==0, -Inf, income_min),
             age_max = ifelse(is.na(age_max), Inf, age_max)) %>% 
      mutate(across(starts_with("income") | starts_with("age"), as.numeric))
}

STE_G17A <- read_csv(STE_paths[1])
STE_G17B <- read_csv(STE_paths[2])
STE_G17C <- read_csv(STE_paths[3])
STE_G17Atidy <- clean_G17(STE_G17A, STE_CODE_2016) 
STE_G17Btidy <- clean_G17(STE_G17B, STE_CODE_2016) 
STE_G17Ctidy <- clean_G17(STE_G17C, STE_CODE_2016) 


STE_G17Atidy %>% head(10) %>% knitr::kable()
```


* Repeat to make `STE_G17B` and `STE_G17C` tidy.
* Combine `STE_G17Atidy`, `STE_G17Btidy` and `STE_G17Ctidy` data by binding rows and save it to the variable `STE_G17tidy`.

<details style = "padding-left:50px;">
<summary>Hint</summary>
```{r, eval = TRUE}
STE_G17tidy <- bind_rows(STE_G17Atidy, 
                         STE_G17Btidy,
                         STE_G17Ctidy)
```
</details>

* Repeat now to get `SA1_G17tidy`.


## `r emo::ji("running")` Exercise 7C

**Quality check your census data**

* According to the data from 7A, how many people are there in Victoria? If you check [http://www.population.net.au](http://www.population.net.au/population-of-victoria/), there are 6.03 million people in 2016 at Victoria. Does it look right? Are we missing some people in the data? If so, what kinds of people are we missing?
* What is the minimum and maximum of values for `count`? Do the range and categories for each column look right?
* Draw a barplot (using `ggplot` or otherwise) of the number people in Victoria by:
  1. sex
  2. age group
  3. income group
  4. sex and age group
  5. sex and income group
  6. age and income group
  7. sex, age and income group

<details style = "padding-left:50px;">
<summary>Hint: click here for a code for the 7th barplot </summary>
  
```{r, eval = TRUE}
STE_G17tidy %>% 
   # remove `P` (should be a combination of female + male)
   filter(sex!="P") %>% 
   # make pretty age and income factors 
   mutate(income = fct_reorder(paste0(scales::dollar(income_min, big.mark = ""), "-", income_max), income_min),
          income = fct_recode(income, 
                              "Nil or negative" = "< $0",
                              "$3000 or more" = "> $3000"),
             age = fct_reorder(paste0(age_min, "-", age_max), age_min),
             age = fct_recode(age, "over 85" = ">85")) %>% 
   # plot
   ggplot(aes(x = age, y = count, fill = sex)) + 
      geom_col(position = "dodge") + 
   facet_wrap(~income) + 
   # let's not use the default theme
   theme_bw(base_size = 12) + 
   theme(axis.text.x = element_text(angle = 90, vjust = 0.3)) + 
   ggtitle("Income, sex & age distribution in Victoria from 2016 Census") 

   
```
  
  
</details>
  
* Do the graphs look right? Discuss with your classmates.



## `r emo::ji("dancing")` Exercise 7D

**Extract statistics from your tidy census data**

According to the 2016 Census data:

* How many women in Victoria are aged between 15-54 years old?
* What is the proportion of people in Victoria that are 25-34 years old (inclusive) and earn $1750 or more per week?
* Suppose I randomly select a man from all the man aged 25-44 years old in Victoria. What is the probability that the man I selected earns less than $1500 per week?



