---
title: 'ETC5512: Instructions for Lab 6'
author: "Lauren Kenendy & Emi Tanaka"
date: "Week 6"
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = TRUE)
set.seed(1)
filter <- dplyr::filter
```


**<span style="color:red">Please do [Preparation](#preparation) *before* the tutorial!</span>**

## `r emo::ji("target")` Objectives

In this tutorial, you will learn to

- interrogate your data by studying summary statistics 
- calculate uncertainties of point estimates


## `r emo::ji("wrench")` Preparation  

   
1. **Set up your project folder**

   Prepare a project calling it say `etc5512-week06` with folders `data` and `analysis` in the project root directory. While you don't have to follow the structure strictly, it's a good habit to maintain a project workflow.
   
   
   
1. **Getting the PISA data**
   * To get the dataset go to  http://www.oecd.org/pisa/data/2018database/
   * Then go under SPSS (TM) Data Files (compressed)
   * And click on Student questionnaire data file 
   * [OPTIONAL] The downloaded file should be called `CY07_STU_QQQ.zip` is about 494MB. Unzip and place the extracted file `CY07_MSU_STU_QQQ.sav` into the `data` folder.
   * Download a smaller version of this data set (about 60MB) [here](https://github.com/numbats/wcd/raw/master/data/pisa2018.rds) and place it into a `data` folder.
   
1. **Install `haven` and `nullabor` R-packages**

```{r, eval = FALSE}
install.packages("haven", "nullabor")
```

4. **Check your project structure**

   Your project structure should look like below.
   
```
etc5512-week08
├── README.md
├── analysis
│   └── exercise.Rmd
├── data
│   ├── CY07_MSU_STU_QQQ.sav [OPTIONAL]
│   └── pisa2018.rds
└── etc5512-week08.Rproj
```



## `r emo::ji("school")` Exercise 6A

**Interrogate the PISA data**

Load the PISA 2018 data in R. You may like to use the smaller data provided in `rds` format instead even if you have the `sav` format one. The `sav` file is over 1GB and so it does take a while to load the data into R. Note however that there are fast ways to import large `csv` files in R (e.g. using `data.table` or `vroom`).



```{r, message = FALSE}
library(tidyverse)
library(haven)
pisa2018 <- readRDS("../lectures/data/lecture-06/pisa2018.rds")
# This is because I had a different repo setup to you to store the data where I was building the lecture slides. 
# Your code will look like:
# pisa2018 <- readRDS(here::here("data","pisa2018.rds")
# OR 
# pisa2018 <- read_csv(here::here("data", "CY07_MSU_STU_QQQ.sav"))
```


(a) How many students participated by each country/region?
(b) Draw a dot plot with mean differences of math scores on the plausible value `PV1MATH` by gender for each country using the weight `W_FSTUWT`.
(c) Now calculate this value with each set of replicate weights (e.g., `W_FSTURWT1` - `W_FSTURWT80`)
(d) Use the formula provided on the slide `Using replicate weights with the PISA` to calculate the variance around this estimate. 
(e)  Add error bars to your plot form (b) based on 95% confidence intervals. 

# What about students who only answered a math component?

(f) How many students conducted the test with `BOOKID` as Form 1-12 or 67-78 (i.e. the number of students who sat for a test that included a maths component) by each country?
(g) Draw the same plot as b. but just based on math scores for students whose test did include a maths component.


## `r emo::ji("student")` Exercise 6B

**What about gender gap in reading and science?**

Draw the same graph as 8A e. using `PV1READ` and `PV1SCIE`. What do you notice?

## `r emo::ji("teacher")` Exercise 6C

**Make better plots**

This question is based on a working book by [Hofmann et al.](https://github.com/heike/data-technologies/blob/master/11-case-study-education.Rmd).

* A problem with the dot plot is that the mean difference axis uses negative numbers, whenever the boys average is higher than the girls. A more accurate reflection on the gender difference would be to use only positive numbers, and label right of 0 to be "girls higher" and left of 0 to be "boys higher". 
* It could also be useful to have a finer resolution of grid lines on the mean difference. 
* `r emo::ji("wrench")` Improve the mean difference axis, so that it has positive numbers only and enhanced labels, and ticks at increments of 5 points.

Add 95% confidence intervals based on the replicate weights. Color the error bar as purple for countries where the girls are significantly performing better than boys and as blue for vice versa.

In other words, your plot should look like below.
```{r, echo = FALSE, fig.height = 12, warning = FALSE, cache = T, message = FALSE}
library(tidyverse)
library(haven)
pisa2018 <- readRDS("../lectures/data/lecture-06/pisa2018.rds") %>% 
   rename(country = CNT, gender = ST004D01T, birth_year = ST003D03T) %>%
  mutate(country = as_factor(country), gender = as_factor(gender))

ncountry <- length(unique(pisa2018$country))

math_df <- pisa2018 %>% 
  mutate(BOOKID = as_factor(BOOKID)) %>%
  filter(BOOKID %in% paste("Form", c(1:12, 67:78))) %>% 
  mutate(gender = as_factor(gender))%>%
  group_by(gender, country) %>% 
  summarise(dplyr::across(c(W_FSTUWT,W_FSTURWT1:W_FSTURWT80), ~ weighted.mean(PV1MATH, w = .x ))) %>%  
  ungroup() %>% 
  pivot_longer(cols = W_FSTURWT1:W_FSTURWT80, 
               names_to = "replicate_ests",
               values_to = "PV1MATH") %>%
  pivot_wider(id_cols = c(country,replicate_ests,W_FSTUWT),
              names_from = gender, 
              values_from = c(PV1MATH, W_FSTUWT)) %>% 
  mutate(rep_female_male_diff = PV1MATH_Female - PV1MATH_Male,
           wtd_female_male_diff = W_FSTUWT_Female - W_FSTUWT_Male)%>%  
  group_by(country,wtd_female_male_diff) %>%
  summarise(math_dif_var = 1/20*sum((rep_female_male_diff - wtd_female_male_diff)^2)) %>%
  ungroup()

math_df %>% 
   mutate(country = fct_reorder(country, wtd_female_male_diff),
          avg = wtd_female_male_diff,
          lower = wtd_female_male_diff - 1.96*sqrt(math_dif_var),
          upper = wtd_female_male_diff + 1.96*sqrt(math_dif_var),
          color = case_when(
          avg > 0 & lower > 0 ~ "purple",
          avg < 0 & upper < 0 ~ "blue",
          TRUE ~ "black"
   )) %>% 
   ggplot(aes(avg, country)) + 
   geom_errorbar(aes(xmin = lower, 
                     xmax = upper,
                     color = I(color))) + 
   geom_point() + 
   geom_vline(xintercept = 0, color = "red") + 
   labs(x = "The difference in mean math scores by gender",
        y = "") + 
   scale_x_continuous(breaks = seq(-20, 20, by = 5),
                      labels = c(seq(20, 0, by = -5), seq(5, 20, by = 5))) + 
   annotate("text", label = c("Boys higher", "Girls higher"), 
            x = c(-10, 10), y = -2) + 
   ggtitle("Gender gap in math scores by country") +
   coord_cartesian(clip = "off", ylim = c(1, ncountry - 1)) + 
   theme(plot.margin = unit(c(1,1,3,1), "lines"),
         axis.title.x = element_text(size = 14, face = "bold", margin = unit(c(3, 0, 0, 0), "lines")),
         plot.title = element_text(size = 18, face = "bold"))
```


