---
title: 'ETC5512: Instructions for Tutorial 5'
author: "Belinda Maher"
date: "Week 5"
output: 
  html_document:
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#devtools::install_github("hadley/emo")
```

## `r emo::ji("target")` Objectives

In this tutorial, you will learn 

- how to load and manipulate data from an Excel file with sheets
- work with tidyverse wrangling verbs to summarise the data and recalculate measures at different levels of aggregation
- practise plotting using ggplot2 to make big picture inferences about the data

## `r emo::ji("wrench")` Data Preparation  

Use this code to process the data from the spreadsheet. Try to understand what the code is doing at each step
```{r load data}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(here))
library(readxl)
library(purrr)
suppressPackageStartupMessages(library(magrittr))
suppressPackageStartupMessages(library(curl))

#Save the file
url<-"https://wcd.numbat.space/lectures/data/lecture-05/Rail%20PT%20Public%20Monthly%20Figures%20cleaned.xlsx"
if(!file.exists(here("Rail PT Public Monthly Figures cleaned.xlsx"))){
curl_download(url, "Rail PT Public Monthly Figures cleaned.xlsx")}

#or adjust this line to where you have saved the file (you might need to add *'folder','subfolder',* depending on where you have saved the file)
filename <- here('Rail PT Public Monthly Figures cleaned.xlsx')
sheets <- excel_sheets(filename)
perf <- map(sheets, ~read_xlsx(filename, sheet=.x))

#Name the sheets in the list by 
names(perf)<-sheets

#Create a list of variables which you want to keep
keep<-c("Parent_Route","Mode","Service_Year", "Service_Date", "Service_Month",
       "Reporting_Group","num_services_scheduled","num_services_cancelled","num_services_short","num_services_bypassed","num_services_scheduled_at_dest","num_services_on_time_at_dest","num_observations_on_time","num_observations_scheduled", "pct_on_time_avg")
#These lines make sure that the data is in the correct format as it is read in from the sheet (not as character).
perf$Yarra$pct_on_time_at_dest%<>% as.numeric()
perf$Yarra$pct_not_on_time_at_dest%<>% as.numeric()
perf$Yarra$pct_on_time_avg%<>% as.numeric()
```


Don't worry about "NAs introduced by coercion" errors the above code raises - this is because not all of the 3 sheets contain all of the variables, so forcing the data to numeric leads to NAs where one of the variables is not present for that mode.


```{r performance}
#stick all 3 sheets together into one data frame from a list
performance <- dplyr::bind_rows(perf, .id="Mode")

#remove extra columns not used (you may want to edit the 'keep' list if you want to look at detailed data at route level)
performance <- dplyr::select(performance,all_of(keep))
```



## 1. `r emo::ji("computer")` Load data and understand the process

<b>1. Question: What does the *.id="Mode"* command do here?</b>


`r emo::ji("eyes")` Have a look at the data:
```{r eval=FALSE}
View(performance)
head(performance)
```

Next, Load this function by running the chunk:
```{r calc_measures}
#the complicated formula which calculates the Punctuality and Reliability measures in different ways for different modes!  
library(dplyr)
  calc_measures<-function(input){input %>% mutate(
  Reliability=case_when(
  Mode =="Metro" ~ ((1-((num_services_cancelled+(0.25*num_services_short)+(0.125*num_services_bypassed))
                      /num_services_scheduled))*100),
  Mode == "Yarra" ~ (1-((num_services_cancelled+ 0.125*num_services_short)
                        /num_services_scheduled))*100,
  Mode == "VLine" ~  (1-(num_services_cancelled/num_services_scheduled))*100),

  Punctuality=case_when(
  Mode =="Metro" ~ (num_services_on_time_at_dest/num_services_scheduled_at_dest)*100,
  Mode == "Yarra" ~ (num_observations_on_time / num_observations_scheduled)*100,
  Mode == "VLine" ~  (num_services_on_time_at_dest/num_services_scheduled_at_dest)*100)
  )
}
```
Once the above function is loaded into your environment, you can use it to calculate without having to repeat that code.
You can now alter the below code to calculate the measures and adjust what to summarise by.

## 2. `r emo::ji("numbers")` Summarising data by different variables


```{r summarise}
library(magrittr)
performance %>% 
  group_by(Service_Date, Service_Year, Service_Month, Mode) %>% 
  summarise(across(starts_with("num_"), sum, .names = "{.col}"),.groups="drop") %>%
  calc_measures() -> perf_month
```

<b>Question: The code above summarises all columns starting with what string?</b>



## 3. Example Code: Summarise by month

```{r eval=FALSE}
head(perf_month)
```


Next step is to keep the id variables and the two performance measures we have calculated, and use pivot_longer to put the data into a tidy format
```{r example_tidy}
perf_month%<>% 
  select("Service_Year","Service_Month", "Mode","Reliability", "Punctuality")  %>%  
  pivot_longer(cols=c(Reliability, Punctuality),names_to="Measure")
```

Have a look at the format of the data now:
```{r eval=FALSE}
head(perf_month)
```


## 4. `r emo::ji("watch")` Average Punctuality - group_by() & summarise()

<b>Question: Identify, using summary functions, which tram route performed the worst (in terms of punctuality) on average in 2020.</b>
Hint: the variable for tram punctuality is called pct_on_time_average
Hint: use dplyr::filter(??==??, ??==??) to filter out the data you need,
then use group_by( *variable* ) %>% summarise( monthly_average = mean( ?? )) %>%
 calc_measures() %>%  arrange(monthly_average)

You could alternatively summarise the data for each Parent_Route by year instead of month, and then compare the values


(note that this is just a quick calculation which doesn't take into account differing numbers of services on different months, but just compares the average by month)


## 5. `r emo::ji("abacus")` Summarise the performance data by Reporting_Group
<b>Exercise</b>
Using the data you have ingested from the Excel file above into *peformance*, summarise by Reporting Group instead of Parent_Route to have a higher level look at the data
(use the example code above in step 3 which shows you how to summarise by month)


## 6.  `r emo::ji("pen")` Punctuality Questions from Reporting_Group data

## `r emo::ji("tram")` Tram Performance by Reporting_group
<b>Question: Which Tram Reporting Group did not perform significantly better in 2020 (on Punctuality)?</b>
(note that the City Shuttle Group has no data because it contains only the City Circle tram which is not measured for on time running)

- Hint: Filter by Tram, and use group = Service_Year, filter the data >=2015 & <=2020
- Hint: Use facet_wrap(~Reporting_Group)
- Hint: Filter for Punctuality values only
- Hint: Use the ggplot2 code shown in the lecture slides to help you

Also don't be afraid to try different options out and see if they work or not!



## `r emo::ji("train")` VLine Performance 2020 compared to previous years
<b>Question: Which Reporting_Group had a month of very poor punctuality in 2020 for V/Line? Alter your code from the previous question</b>

- Hint: Filter by VLine, then as before:
- use group = Service_Year, filter the data >=2015 
- Hint: Use facet_wrap(~Reporting_Group)
- Hint: Filter for Punctuality values only








