---
title: 'ETC5512: Instructions for Lab 10'
author: "Will Mackey"
date: "Week 10"
output: 
  html_document:
    toc: true
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(eval = TRUE, warning = FALSE, message = FALSE)
set.seed(1)
filter <- dplyr::filter
```


**<span style="color:red">Please do [Preparation](#preparation) *before* the tutorial!</span>**

## `r emo::ji("target")` Objectives

In this tutorial, you will learn to

- interrogate your data by studying summary statistics 
- calculate uncertainties of point estimates


## `r emo::ji("wrench")` Preparation  

1. **Read** Sections 13.1-13.4 of [_R for Data Science_](https://r4ds.had.co.nz/relational-data.html)
   
1. **Set up your project folder**

 Prepare a project calling it say `etc5512-week10` with folders `data` and `analysis` in the project root directory. While you don't have to follow the structure strictly, it's a good habit to maintain a project workflow.

1. **Install `tidyverse`**

```{r, eval = FALSE}
install.packages("tidyverse")
```

2. **Check your project structure**

   Your project structure should look like below.
   
```
etc5512-week10
├── README.md
├── analysis
│   └── exercise.Rmd
├── data
└── etc5512-week10.Rproj
```

3. Load the relevant packages

```{r exercise}
library(tidyverse)
library(readxl)
library(lubridate)
```



## `r emo::ji("banknote")` Exercise 10A

**Is there a difference between female and male pay for the same hours in the same occupation?**

* Download detailed income data from the ATO and read into R.

```{r q1}
ato_url <- "https://data.gov.au/data/dataset/23b8c299-a85b-4fc0-a07d-5ed14e23a103/resource/6a989b47-29ce-451a-a9ef-ed57be644b61/download/ts18individual14occupationsextaxableincomerange.xlsx"
```

* For each year, sex and occupation, find the average salary or wage for people who had a salary or wage. 

```{r s1}
if (!file.exists(here::here("data", "ato-pit.xlsx"))) {
  download.file(ato_url, here::here("data", "ato-pit.xlsx"))
}

## For each financial year, sex and occupation, find the average 'Salary or wages'
ato_raw <- read_excel(here::here("data", "ato-pit.xlsx"), 
                      sheet = "Individual Table 14B", 
                      skip = 3) %>% 
  select(fy = 1,
         ato_occ = 2,
         sex = 3,
         n = 6, 
         wages = 7)

ato_summary <- ato_raw %>% 
  group_by(fy, ato_occ, sex) %>% 
  summarise(ato_n = sum(n),
            ato_wages = sum(wages),
            ato_average_wage = ato_wages / ato_n)
```


* Download detailed hours data from the ABS: 

```{r q2}
abs_working_url <- "https://www.abs.gov.au/statistics/labour/employment-and-unemployment/labour-force-australia-detailed/mar-2021/EQ08.xlsx"

```

* For each year, sex and occupation, find the average number of hours worked

```{r s2}

if (!file.exists(here::here("data", "abs-hours.xlsx"))) {
  download.file(abs_working_url, here::here("data", "abs-hours.xlsx"))
}


## For each financial year, sex and occupation, find the average number of hours worked
abs_raw <- read_excel(here::here("data", "abs-hours.xlsx"), 
                      sheet = "Data 1", 
                      skip = 3) %>% 
  select(date = 1,
         abs_occ = 4,
         sex = 2,
         state = 3,
         n_000 = 5, 
         hours_000 = 6)

abs_clean <- abs_raw %>%
    mutate(n = n_000 * 1000,
           hours = hours_000 * 1000,
           year = year(date),
           quarter = quarter(date, with_year = TRUE, fiscal_start = 7),
           fy = substr(quarter, 1, 4))
  
abs_summary <- abs_clean %>% 
  group_by(fy, abs_occ, sex) %>% 
  summarise(abs_n = sum(n),
            abs_hours = sum(hours),
            abs_average_hours = abs_hours / abs_n)

```


* Before joining the datasets by `year`, `sex` and `occupation`: 
explore and reconcile the differences in values in these two datasets.

```{r s3}
# Need to join on sex, fy, and occupation.
# compare these variables:
abs_summary$sex %>% unique()
ato_summary$sex %>% unique()

abs_summary$fy %>% unique()
ato_summary$fy %>% unique()

abs_summary$abs_occ %>% unique() %>% .[1:10]
ato_summary$ato_occ %>% unique() %>% .[1:10]

# The occupation code looks to be our best bet here. Extract it:
abs_join <- abs_summary %>% 
  mutate(occ_code = substr(abs_occ, 1, 4))

ato_join <- ato_summary %>% 
  mutate(occ_code = substr(ato_occ, 1, 4))

# Represent the financial year with just the end of the year (eg 2010-2011 -> 2011)
ato_join <- ato_join %>% 
  mutate(fy = paste0("20", substr(fy, 6, 7)))

# Remove wayward "i. all ranges" category from ato, and add
ato_join <- ato_join %>% 
  filter(sex != "i. all ranges") %>% 
  mutate(sex = paste0(sex, "s"))
```

* Join datasets to find average wage per hour worked by occupation and sex

```{r s4}
# All set!
wages_hours <- ato_join %>% 
  left_join(abs_join)
```

* Conduct some post-join checks:
  * does the resulting joined dataset have the same number of observations as the ATO dataset used in the join? Why/why not?
  * does the resulting joined dataset have the same number of observations as the ABS dataset used in the join? Why/why not?
  * how could you tell which of the ATO observations weren't matched with an ABS observation? 
  * what number/% of observations weren't matched?
  * what were the ato_join occupations that did not match?
  * why weren't they matched?

```{r s5}
# What variables were used to join the two datasets?
# Joining, by = c("fy", "occ_code")

# Does the wages_hours dataset have the same number of observations
# as the ato_join dataset? Why/why not?
nrow(wages_hours)
nrow(ato_join)

# Does the wages_hours dataset have the same number of observations
# as the abs_join dataset? Why/why not?
nrow(wages_hours)
nrow(abs_join)


# how could you tell which of the ato_join observations weren't matched with
# an abs_join observation? What number/% of observations weren't matched?
wages_hours %>% 
  ungroup() %>% 
  summarise(count = n(),
            count_no_join = sum(is.na(abs_occ)),
            pc_no_join = count_no_join / count)


# What were the ato_join occupations that did not match?
wages_hours %>% 
  filter(is.na(abs_occ)) %>% 
  pull(ato_occ) %>% 
  unique()

# Three groups of not matched:
# 1. 'type not specified'; codes ending in 00
# 2. ATO-specific codes (outside of ANZSCO); codes ending 
# 3. small wages

```


* Generate a new variable `wage_rate`, which is average income divided by average wages.

* Explore:
  * average difference in `wage_rate` for females and males in the same occupation. 
  * plot the 20 occupations with the highest and lowest difference in `wage_rate` between females and males.

```{r s6}
wages_hours <- wages_hours %>% 
  select(fy, occ_code, ato_occ, sex,  ato_n, ato_average_wage, abs_average_hours) %>% 
  mutate(ato_weekly_wage = ato_average_wage / 52,
         wage_rate = ato_weekly_wage / abs_average_hours)

# Plot a histogram of wage_rates
wages_hours %>% 
  ggplot(aes(wage_rate)) +
  geom_histogram()  +
  facet_grid(sex~.) + 
  scale_x_continuous(limits = c(0, 100))

# Calculate the dollar and percent difference between male and female wages
wages_hours <- wages_hours %>% 
  arrange(fy, occ_code, ato_occ, sex) %>% 
  group_by(fy, occ_code, ato_occ) %>% 
  mutate(female_diff_dollar = wage_rate - lead(wage_rate),
         female_diff_pc = wage_rate / lead(wage_rate) - 1)

# What is the mean of female_diff_pc?
# hint: use weighted.mean, with ato_n as the weights
female_diff <- wages_hours %>% 
  filter(female_diff_pc < Inf) %>% 
  group_by(fy) %>% 
  summarise(female_diff_pc = weighted.mean(female_diff_pc, ato_n, na.rm = TRUE))

# Plot the average difference between female and male wages over time
female_diff %>% 
  ggplot(aes(fy, female_diff_pc)) + 
  geom_col()
```



## `r emo::ji("party")` Exercise 10B

**Better safe than sorry**

* Write a function called `safe_left_join`, which performs as `left_join(a, b)` does,
but checks that the resulting dataframe has the same number of observations as 
dataset `a`. 

```{r s7}
safe_left_join <- function(a, b, ..., warn = TRUE) {
  
  joined <- left_join(a, b, ...)
  
  if (warn & nrow(joined) != nrow(a)) {
    warning("The join created more observations")
  }
  
}
```



