---
title: 'ETC5512: Instructions for Tutorial 2'
author: "Lecturer: Kate Saunders"
date: "Week 2"
output: 
  html_document:
    css: tutorial.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      eval = FALSE,
                      cache = FALSE, 
                      warning = FALSE, 
                      message = FALSE)
options(width=80,digits=3)
```  
    
  
# `r emo::ji("target")` Objectives
  
In this tutorial, you will learn 

- Identify whether data are experimental or observational
- Delineate the data collection methods
- Logically suggest the population that a sample represents.  
- Using and accessing open data sources

# `r emo::ji("wrench")` Preparation

1. The install the following R packages. (You can copy and paste the code below into your RStudio Console window to do this.):

```
install.packages(c("lubridate", "ggthemes", "forcats"))
```

and if you missed tutorial 1, you will also need to install packages listed in those instructions.

2. Complete the assigned reading, and the associated quiz (in flux). 

# `r emo::ji("waving_hand")` Getting started

Say hello to your tutor and to your neighbours! 

# Tutorial


## `r emo::ji("gear")` Exercise 1 

<!-- observational or experiment, and according to my taxonomy, describe types of variables, brainstorm questions, given a question define population-->

This question relates to the [Tidy Tuesday Data on locations of alternative fuel recharging stations](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-03-01/readme.md). Have a read through this site, and also visit the link to the data providers, DOT.

### a. Read the details about the data at [DOT](https://data-usdot.opendata.arcgis.com/datasets/usdot::alternative-fueling-stations/about). How is this data collected, do you think? 

### b. What type of data is this? (observational, experimental, survey, census)

### c. Describe the population, and what is the sample. 

### d. Download the data and plot the fueling locations on a map, coloured by fuel type.

```{r}
library(tidyverse)
library(ggthemes)
library(lubridate)
stations <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-03-01/stations.csv')
# Filter to continental USA, but this cannot 
# be done using states because some records have 
# erroneous lat/long
# stations <- stations %>%
#  filter(!(STATE %in% c("AK", "HI", "PR", "ON")))

# Map the sites
usa <- map_data("state")
# Filter to continental USA using map boundary
stations <- stations %>%
  filter(between(LONGITUDE, min(usa$long)-1, max(usa$long)+1),
         between(LATITUDE, min(usa$lat)-1, max(usa$lat)+1))
ggplot() + 
  geom_path(data=usa, aes(x=long, y=lat, group=group), colour="grey80") + 
  geom_point(data=stations, aes(x=LONGITUDE, 
                                y=LATITUDE,
                                colour=FUEL_TYPE_CODE),
             alpha=0.8) +
  facet_wrap(~FUEL_TYPE_CODE, ncol=4) +
  coord_map() +
  theme_map() +
  theme(legend.position = "none")
```

### e. Count the number of new stations by month, and make a time series plot by fuel type. 

```{r}
# Time line of opening
stations %>% 
  mutate(OPEN_DATE = as.Date(OPEN_DATE)) %>%
  filter(!is.na(OPEN_DATE)) %>%
  mutate(m = month(OPEN_DATE),
         yr = year(OPEN_DATE)) %>%
  #filter(y < 2022) %>%
  mutate(open_yrmth = as.Date(paste(yr, m, "01", sep="-"), "%Y-%M-%d")) %>% 
  group_by(open_yrmth, FUEL_TYPE_CODE) %>%
  summarise(nopen = n(), .groups = "drop") %>%
ggplot(aes(x=open_yrmth, 
           y=nopen, 
           colour=FUEL_TYPE_CODE)) +
  geom_line() +
  facet_wrap(~FUEL_TYPE_CODE, ncol=4, scales="free_y") +
  ylab("# opening") +
  scale_x_date("", date_labels="%y") +
  theme(legend.position = "none")
```

### g. If the question to answer is "which alternative fuel vehicle is the fastest growing?" what is the explanatory (independent, predictor) variable and what is the response variable?


## `r emo::ji("gear")` Exercise 2

Here we will look at the [Chocolate bar ratings](https://github.com/rfordatascience/tidytuesday/blob/master/data/2022/2022-01-18/readme.md). Details (brief) of how the data was collected are provided [here](https://github.com/rfordatascience/tidytuesday/issues/408) and more about the data itself is [here](http://flavorsofcacao.com/chocolate_database.html).

```{r}
chocolate <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2022/2022-01-18/chocolate.csv')
```

### a. What type of data is this? (observational, experimental, survey, census)

### b. How is the data collected?

### c. Describe the population.

### d. For the question "Which country of origin of the bean obtains the best rating?" state the response and predictor variables.

### e. Make a plot to answer the previous question. (Only use countries with more than 10 records.)

```{r}
library(forcats)
keep <- chocolate %>% 
  count(country_of_bean_origin, sort = TRUE) %>%
  filter(n>10) %>%
  pull(country_of_bean_origin)
chocolate %>%
  filter(country_of_bean_origin %in% keep) %>%
  ggplot(aes(x=fct_reorder(country_of_bean_origin, rating, mean), 
             y=rating)) +
    geom_jitter(width=0.1) + 
    stat_summary(fun = mean, fun.min = median, fun.max = median,
                 geom = "point", colour = "orange") +
    xlab("") + 
    coord_flip()
```


## `r emo::ji("gear")` Exercise 3

Read the description of the study titled ["Clearing the Fog: Is Hydroxychloroquine Effective in Reducing COVID-19 Progression (COVID-19)"](https://clinicaltrials.gov/ct2/show/NCT04491994). 

### a. What type of data is this? (observational, experimental, survey, census)

### b. How many subjects participated in the study at the start, and to completion?


### c.What are the:

- experimental units? 
- factor? 
- blocking factors?
- response variable (outcome measures)?

### d. How are subjects assigned to treatment groups?

### e. What were the results of the study?

### f. Construct the data from the results reported. Compute the proportion of subjects with progression of COVID after 5 days, for the two treatments. Include the standard error of the estimate.

```{r}
hcq <- tibble(trt = c("standard", "standard", "hcq", "hcq"),
              progression = c("all", "yes", "all", "yes"), 
              count = c(151, 5, 349, 11))
hcq %>% 
  pivot_wider(names_from = "progression", values_from = "count") %>%
  mutate(p = yes/all) %>%
  mutate(se = sqrt(p*(1-p)/all))
```

### g. Based on the proportions and their standard errors, why would the result of the study be that HCQ does NOT improve the outcomes of COVID patients?

### h. What is the population for this experiment?

## Optional extra learning exercise

This tutorial will prepare you for the material of next week. Please follow the following steps:

1. Enter the address https://olc.worldbank.org/ in your browser.

2. Click on "Register" and create a profile.

3. Once you login, the following link can be used to enroll in the course Open Data for Data Users (Self-Paced), 
https://olc.worldbank.org/content/open-data-data-users-self-paced

4. Please allow the pop up blocker at the browser settings to run the course successfully.

5. Follow the whole course and answer the questions throughout.

# `r emo::ji("stop_sign")` Wrapping up

Talk to your tutor about what you think you learned today, what was easy, what was fun, what you found hard.

##### Material originally developed by Prof. Di Cook and maintained by Dr. Kate Saunders
##### © Copyright 2024 Monash University
