---
title: 'ETC5512: Instructions for Lab 3'
author: "Didier Nibbering"
date: "Week 3"
output: html_document
---


# Objectives
  
In this tutorial, you will learn 

- to download macroeconomic data
- assess the collection methods and the quality of the data
- write computer code to wrangle and analyse data


# Exercise 3A: Survey of Professional forecasters

Download the individual responses for the Survey of Processional forecasters for U.S. nominal gross domestic product. The Survey of Processional forecasters can be found here: https://www.philadelphiafed.org/research-and-data/real-time-center/survey-of-professional-forecasters

- Select "Download one variable at a time"
- Select "Nominal Gross National Product/Gross Domestic Product (NGDP)". Note that you can find "Documentation.pdf" here. 
- Select "Individual Responses"
- This will download the file "Individual_NGDP.xlsx"

Use the lecture materials and the Survey documentation to answer the following questions:

1. What is the source for the real-time  historical values for nominal gross domestic product?
2. Explain the difference between the data prior to 1992, and after 1992.
3. Why is the data for nominal gross domestic product seasonally adjusted?
4. Is nominal gross domestic product a stock or a flow variable? 
5. Discuss whether nominal gross domestic product is a good measure for growth of production.
6. Consider the forecasts for nominal gross domestic product for the first quarter of 2005. What was the deadline date for submitting the forecasts?

Use the downloaded file "Individual_NGDP.xlsx" to answer the following questions:

7. Put the downloaded file in your working directory and use read_excel to load the data in your workspace
```{r A7, eval = FALSE}
library(tidyverse)
library(readxl)
spf <- read_excel(???,sheet=1)
```
8. Select the forecasts (nowcasts) for nominal gross domestic product for the first quarter of 2005.
```{r A8, eval=FALSE}
spf2005Q1for <- spf %>% filter(YEAR==???,QUARTER==???) %>% 
  mutate(forecast=as.double(???)) %>% 
  select(???,???) %>% drop_na()  
```  
9. What is the mean forecast for the first quarter of 2005?
10. Who is the most optimistic forecaster?
  
  
# Exercise 3B: Real-time data vintages
  
Use the ALFRED package to access the ALFRED database. Download all the vintages for the nominal gross domestic product in the first quarter of 2005.
```{r B,eval=FALSE}
library(alfred)
fred2005Q1vin <- as_tibble(get_alfred_series(???, 
                                             observation_start = ???,
                                             observation_end = ???,
                                 realtime_start = ???, 
                                 realtime_end = ???))
```

1. What is the source of the ALFRED nominal gross domestic product data?
2. Provide the license of the ALFRED data.
3. What is the earliest vintage in ALFRED after the survey submission of the Survey of Professional Forecasters?
4. Plot the values for nominal gross domestic product in the different vintages over time.
```{r B4,eval=FALSE}
fred2005Q1vin %>% ggplot(aes(y=???,x=???))+geom_line()
```
5. How can the observed pattern be explained?
  
  

