---
title: 'ETC5512: Solutions for Lab 6'
author: "Patricia Menéndez"
output:
  html_document: default
keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.path = "Tutorial5_Images/", cache = TRUE, fig.width = 8, fig.height = 8)

```

# Reading libraries

```{r , message = FALSE, warning = FALSE}
library(tidyverse)
library(leaflet)
library(ggmap)
library(readr)
library(mapview)
library(viridis)
library(rgdal)
library(lubridate)

```

# Tutorial Objectives

- Explore LTMP data measurements for  hard coral cover and algae cover. 
- Visualize variability in mean estimates.
- Run a simple regression model.
- Practice your data wrangling skills.




# Exercise 2: Create a report and read the data

- Create an Rstudio project.
- Inside your Rstudio project folder create a reproducible Rmd file called *tutorial6.Rmd*. The file should  render into an html file.
- Add the *Data* folder place the LTMP data inside that folder.
- Read the LTMP data.
```{r , message = FALSE, warning = FALSE}
# LTMP data
ltmp <- read_csv(file = "Data/ltmp_hc_sc_a_by_site/ltmp_hc_sc_a_by_site.csv", col_types = cols())


```
- Add an R code chunk to include the relevant R libraries at the top of the *tutorial6.Rmd* file.
- Prepare the data so that you can create the following figures and discuss the differences between those graphs:

**Hint:**
```{r, eval = FALSE}
ltmp %>% dplyr::filter(GROUP_CODE == "Hard Coral",
                           SECTOR == XX
                           ) %>%
                    mutate(REEF_NAME = as.factor(REEF_NAME),
                    Year = year(XX)) %>%
                   group_by(SECTOR, SHELF, Year) %>%
  summarize_at(vars(XX), list(covermean = XX, coversd = XX)) %>%
  mutate(samplesize = n()) %>%
  mutate(upper = covermean + coversd/sqrt(samplesize),
         lower = covermean - coversd/sqrt(samplesize))
```



```{r , echo = TRUE}

plot1dat = ltmp %>% dplyr::filter(GROUP_CODE == "Hard Coral",
                           SECTOR == "CA",
                           SITE_NO %in% c("1", "2", "3")
                           ) %>%
  mutate(REEF_NAME = as.factor(REEF_NAME),
         Year = year(SAMPLE_DATE)) %>%
  group_by(SECTOR, SHELF, Year) %>%
  summarize_at(vars(COVER), list(covermean = mean, coversd = sd)) %>%
  mutate(samplesize = n()) %>%
  mutate(upper = covermean + coversd/sqrt(samplesize),
         lower = covermean - coversd/sqrt(samplesize))



ggplot(plot1dat, aes( x = Year, y = covermean)) +
  geom_line() +
  geom_pointrange(aes(ymin=lower, ymax=upper)) +
  facet_wrap(~SHELF, ncol = 1)
```

**Hint:**


```{r , eval = FALSE}
 ltmp %>% dplyr::filter(GROUP_CODE == XX,
                           SECTOR == "CA") %>%
  mutate(REEF_NAME = as.factor(REEF_NAME),
         Year = year(SAMPLE_DATE)) %>%
  group_by(XX, XX, XX) %>%
  summarize_at(vars(COVER), list(covermean = mean, coversd = sd)) %>%
  mutate(upper = covermean + XX,
         lower = covermean - XX) 

```


```{r , echo = TRUE}

plot2dat = ltmp %>% dplyr::filter(GROUP_CODE == "Hard Coral",
                           SECTOR == "CA",
                           SITE_NO %in% c("1", "2", "3")
                           ) %>%
  mutate(REEF_NAME = as.factor(REEF_NAME),
         Year = year(SAMPLE_DATE)) %>%
  group_by(SECTOR, SHELF, Year) %>%
  summarize_at(vars(COVER), list(covermean = mean, coversd = sd)) %>%
  mutate(upper = covermean + coversd,
         lower = covermean - coversd) 


ggplot(plot2dat, aes( x = Year, y = covermean)) +
  geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "blue", alpha = 0.4) +
  facet_wrap(~SHELF, ncol = 1)
```

# Exercise 3: Data wrangling to create summary variables

- Filter the LTMP data set using the variables *GROUP_CODE* and *REEF_NAME* so that you only have values for *Hard Coral* and *Algae*  for 
the *LOW ISLANDS REEF*.
- Create a new  variable called *Year* and extract the *year* value from the *SAMPLE_DATE* variable.
- Group the data set by *Year* and *GROUP_CODE* and compute for those groups the mean and stand deviation for the *COVER* variable.  Your data should look like this:


```{r, eval = TRUE, echo = TRUE}
regdata = ltmp %>% dplyr::filter(GROUP_CODE %in% c("Hard Coral", "Algae"),
                       REEF_NAME == "FITZROY ISLAND REEF") %>%
  mutate(Year = year(SAMPLE_DATE)) %>%
  group_by(Year, GROUP_CODE) %>%
  summarize_at(vars(COVER), list(covermean = mean, coversd = sd), na.rm = TRUE) 

head(regdata)

```

- Convert your data into wide format so that you have two new columns where *Hard Coral* and *Algae* cover values are stored. Your data should look like this:

```{r , echo = TRUE}
regdataWF = ltmp %>% dplyr::filter(GROUP_CODE %in% c("Hard Coral", "Algae"),
                       REEF_NAME == "LOW ISLANDS REEF") %>% 
  mutate(Year = year(SAMPLE_DATE)) %>%
  group_by(Year, GROUP_CODE) %>%
  summarize_at(vars(COVER), list(covermean = mean), na.rm = TRUE) %>%
  pivot_wider(names_from = GROUP_CODE, values_from = covermean) 


head(regdataWF)
```
- Plot your newly created data to produce the following figure:

```{r }

ggplot(regdataWF, aes(x = `Hard Coral`, y = Algae)) +
  geom_point() 
```

# Exercise 4: Practicing liner regression to understand the relationship between hard coral cover and algae abundance.

- Using the data that you created in Exercise 3 (wide format data). 
- Run a linear regression model to understand the relationship between hard coral cover and algae.
- Add two new variables into the data set to store your predicted and residuals values from the regression.



```{r  , echo = TRUE}

reg = lm(Algae ~ `Hard Coral`, data = regdataWF )
```


```{r , echo = TRUE}
regdataWF$predicted <- predict(reg)
regdataWF$residuals <- residuals(reg)
```
- Create the following figure and change the colours: 

```{r , echo = TRUE}
ggplot(regdataWF, aes(x = `Hard Coral`, y = Algae)) +
  geom_smooth(method = "lm", se = FALSE, color = "lightblue") +
  geom_segment(aes(xend = `Hard Coral`, yend = predicted), alpha = .2) +
  geom_point(aes(color = abs(residuals), size = abs(residuals))) + 
  scale_color_continuous(low = "#D19494", high = "#D92626") +
  guides(color = FALSE, size = FALSE) +  # Size legend also removed
  geom_point(aes(y = predicted), shape = 1) +
  theme_bw() +
  ggtitle("Linear regression model")
```





