---
title: 'ETC5512: Instruction to Open Data'
author: "Lecturer: Kate Saunders"
format: 
  html:
    toc: true
    css: ../../assets/tutorial.css
    embed-resources: true
  pdf:
    toc: true
editor: visual
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Learning Objectives

-   Learn your way around RStudio.
-   Become confident doing basic computation using R in RStudio.

### Before your tutorial

**Ensure you have installed both R and RStudio.**

- Follow theses [installation instructions](https://startr.numbat.space/) 

Also and work through the following [startR modules](https://startr.numbat.space/):

-   Do the module on RStudio Basics (Module 2).

-   Do the module on R Basics (Module 3).

These should take you \~ 40 minutes.

## Package Installation 

Run the below code to install the packages we need for today's tutorial. Remember you only need to install packages once.

```{r}
#| eval: false
install.packages("tidyverse")
install.packages("galah")
```

Once the packages are installed load in the libraries so we can use them in today's class. 

```{r}
#| eval: false
library(tidyverse)
library(galah)
```

Installing packages is like screwing in a lightbulb, loading the library is like turning the switch on and off. You only need to screw in the lightbulb once (install the package), but you can turn the light on many times (load the library).

## Exercise 1: Atlas of Living Australia

The Atlas of Living Australia is a major resource for occurrence data on animals, plants, insects, fish.

### a. Download

i. Point your browser to https://www.ala.org.au. Check the terms of use. Does it have a license?

ii. Using the `galah` library, and the function `occurrences` extract the records for platypus. To download the data from this API **you will need to register with your email** first. Once you've done that change the code below so it has your email.

```{r}
#| eval: false 

library(galah)

# Change the below code to your email address
your_email = "ADD_YOUR_EMAIL_HERE"

galah_config(email = your_email,
             download_reason_id = 10, 
             verbose = TRUE)

platypus <- galah_call() |> 
  galah_identify("Ornithorhynchus anatinus") |> 
  atlas_occurrences()
```

Take a look at the data you downloaded. 

```{r}
#| eval: false
View(platypus)
```

iii. Let's do some data wrangling to tidy up our data. 

```{r}
#| eval: false
platypus <- platypus |> 
  rename(Longitude = decimalLongitude,
         Latitude = decimalLatitude) |>
  mutate(eventDate = as.Date(eventDate)) |>
  filter(!is.na(eventDate)) |>
  filter(!is.na(Longitude)) |>
  filter(!is.na(Latitude))
```

iv. Save our result. To save both your data and the file your are working on in the same place, in the top menu go to Session \> Set Working Directory \> To Source File Location. (If you are already familiar with R Projects you may like to use that instead.)

```{r}
#| eval: false
platypus_file_name = "platypus.csv" 
write_csv(platypus, file = platypus_file_name)
```

### b. Data quality checks

i. Plot the locations of sightings. Where in Australia are platypus found?

```{r}
#| eval: false 

read_csv(platypus_file_name)

ggplot(platypus, aes(x=Longitude, y=Latitude)) +
  geom_point()
```

ii. What dates of sightings are downloaded?

```{r}
#| eval: false
platypus |> select(eventDate) |> summary()     
```
### c. Data collection methods

How is this data collected? Explain the ways that a platypus sighting would be added to the database. Also think about what might be missing from the data?

## Learning to code 

:::callout-tip

When you are getting started if can be useful check your basic understanding of what each piece of code does. 

Some tips: 

* Use the help menu to look up what functions do, e.g. run `?write_csv`

* Copy and paste code chunks in Generative AI and ask it to explain to you through what the code is doing.

It can then be useful to go back and add comments to your code to explain what the different pieces do.

:::

## Additional Exercises 

i. Try finding examples of each of the different flavours of data we discussed in class.

* Find the license information

* Find the meta data. 

ii. Also be sure to complete the Open Data Insitute Quizzes from Lectures to consolidate your understanding.

##### © Copyright Monash University
